---
title: "Analysis"
subtitle: "Nominal Surge, Real Stagnation, and the Affordability Crisis 2010-2024"
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(scales)
library(knitr)
library(tidyr)
library(forcats)
library(stringr)
library(DT)

load("data/housing_market_tr.RData")

pal <- list(
  primary   = "#1a1814",
  secondary = "#2c3e50",
  accent    = "#c0392b",
  good      = "#27ae60",
  warn      = "#e67e22",
  muted     = "#95a5a6"
)

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title       = element_text(face = "bold", size = 13, hjust = 0),
      plot.subtitle    = element_text(colour = "#6b6560", size = 10.5, hjust = 0),
      plot.caption     = element_text(colour = "#95a5a6", size = 8, hjust = 0),
      plot.background  = element_rect(fill = "#f8f5f0", colour = NA),
      panel.background = element_rect(fill = "#f8f5f0", colour = NA),
      panel.grid.major = element_line(colour = "#e0dbd3"),
      panel.grid.minor = element_blank(),
      legend.position  = "top",
      legend.text      = element_text(size = 9),
      axis.text        = element_text(colour = "#6b6560", size = 9),
      axis.title       = element_text(colour = "#1a1814", size = 10)
    )
)
```

::: {.callout-note icon="false"}
### Key Takeaways

- **Nominal surge, real stagnation.** KFE rose ~4,800% nominally (2010-2024);
  in real terms the index peaked at 153.7 in 2022 and fell to 118.4 by 2024.
- **Affordability collapsed.** Months of minimum wage to buy 100 m2 rose from
  ~110 in 2010 to ~276 in 2022 -- a 2.5x worsening.
- **Exchange rate pass-through is strong.** Quarterly USD/TRY and KFE changes
  are strongly correlated; TRY depreciation episodes directly precede KFE spikes.
- **Istanbul and Aegean lead, Southeast lags.** Five-year cumulative nominal
  gains (2020-2024) range from ~2,100% to ~1,400% across 26 NUTS-2 regions.
- **Supply is contracting.** Building permits fell from 1.22 million units in
  2015 to 549,000 in 2024 -- a 55% drop.
:::

---

## 1. Nominal vs. Real Price Evolution

### 1.1 The two stories of Turkish housing prices

```{r nominal-vs-real, fig.height=4.5}
annual_macro |>
  select(year,
         `Nominal KFE` = kfe_nominal_2010base,
         `Real KFE`    = kfe_real_2010base) |>
  pivot_longer(-year, names_to = "series", values_to = "value") |>
  ggplot(aes(year, value, colour = series, linetype = series)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_colour_manual(values = c(pal$accent, pal$secondary), name = NULL) +
  scale_linetype_manual(values = c("solid", "dashed"), name = NULL) +
  scale_x_continuous(breaks = 2010:2024) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "Nominal vs. real KFE, 2010-2024 (2010 = 100)",
    subtitle = "Nominal growth is dramatic; real gains are modest and recently reversed",
    x = NULL, y = "Index (2010 = 100)",
    caption  = "Source: TCMB KFE; deflated using TURKSTAT CPI."
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The nominal index rose ~48x from 2010 to 2024. The real index peaked at 153.7
in 2022 -- just 54% above its 2010 level -- before declining. By 2024, real
house prices were lower than in 2022, meaning buyers who purchased at the peak
paid a substantial inflation premium with no real asset gain.

### 1.2 Annual change rates: KFE vs. CPI

```{r yoy-change, fig.height=4}
annual_macro |>
  filter(!is.na(kfe_nominal_yoy)) |>
  select(year,
         `KFE nominal YoY %` = kfe_nominal_yoy,
         `CPI YoY %`         = cpi_yoy_pct,
         `KFE real YoY %`    = kfe_real_yoy) |>
  pivot_longer(-year, names_to = "series", values_to = "pct") |>
  ggplot(aes(year, pct, fill = series)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_hline(yintercept = 0, colour = pal$primary, linewidth = 0.4) +
  scale_fill_manual(
    values = c(pal$accent, pal$muted, pal$secondary),
    name   = NULL
  ) +
  scale_x_continuous(breaks = 2011:2024) +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  labs(
    title    = "Annual price change: KFE vs. CPI, 2011-2024",
    subtitle = "Real KFE turned negative in 2018 and again from 2023",
    x = NULL, y = "Year-on-year change (%)",
    caption  = "Source: TCMB KFE, TURKSTAT CPI."
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The real KFE went negative in two episodes: 2018-2019 (currency crisis) and
from 2023 onwards (inflation ran ahead of price growth). In 2022 the nominal
KFE surged ~183% YoY, yet CPI at ~72% still limited real gains to ~111%.

### 1.3 New housing vs. existing housing

```{r new-vs-existing, fig.height=4}
kfe_new_exist |>
  select(year,
         `All homes`      = kfe_all_2010base,
         `New homes`      = kfe_new_2010base,
         `Existing homes` = kfe_existing_2010base) |>
  pivot_longer(-year, names_to = "series", values_to = "value") |>
  ggplot(aes(year, value, colour = series)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_colour_manual(
    values = c(pal$muted, pal$secondary, pal$accent),
    name   = NULL
  ) +
  scale_x_continuous(breaks = 2010:2024) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "KFE by housing type, 2010-2024 (2010 = 100)",
    subtitle = "New homes (YKFE) consistently outpace existing homes (YOKFE)",
    x = NULL, y = "Index (2010 = 100)",
    caption  = "Source: TCMB KFE -- YKFE (new) and YOKFE (existing)."
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### 1.4 Cumulative growth: housing vs. inflation

```{r cumulative, fig.height=4}
annual_macro |>
  mutate(
    cpi_cumul = cpi_yoy_pct,
    kfe_cumul = kfe_nominal_2010base / 100
  ) |>
  select(year, `Nominal KFE` = kfe_cumul) |>
  pivot_longer(-year) |>
  ggplot(aes(year, value, colour = name)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_colour_manual(values = c(pal$accent), name = NULL) +
  scale_x_continuous(breaks = 2010:2024) +
  scale_y_continuous(labels = label_number(suffix = "x")) +
  labs(
    title    = "Cumulative KFE multiplier since 2010",
    subtitle = "Nominal KFE is 48x its 2010 value by 2024",
    x = NULL, y = "Cumulative multiplier (2010 = 1x)",
    caption  = "Source: TCMB KFE."
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### 1.5 Three major cities vs. national

```{r three-cities, fig.height=4}
city_monthly |>
  select(date, year,
         Istanbul = kfe_istanbul,
         Ankara   = kfe_ankara,
         Izmir    = kfe_izmir,
         Turkey   = kfe_turkey) |>
  pivot_longer(-c(date, year), names_to = "city", values_to = "kfe") |>
  ggplot(aes(date, kfe, colour = city)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_colour_manual(
    values = c(pal$accent, pal$good, pal$secondary, pal$muted),
    name   = NULL
  ) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "KFE by major city vs. national average (2010 = 100)",
    subtitle = "Istanbul and Izmir consistently outperform the national index",
    x = NULL, y = "Index (2010 = 100)",
    caption  = "Source: TCMB KFE (selected reference periods 2020-2024)."
  )
```

---

## 2. Regional Analysis (26 NUTS-2 Regions)

### 2.1 Five-year cumulative change (2020-2024)

```{r regional-bar, fig.height=7}
regional_nuts2 |>
  filter(!is.na(cum_2020_2024_pct)) |>
  mutate(region_name = fct_reorder(region_name, cum_2020_2024_pct)) |>
  ggplot(aes(cum_2020_2024_pct, region_name)) +
  geom_col(fill = pal$secondary) +
  geom_text(
    aes(label = paste0(round(cum_2020_2024_pct, 0), "%")),
    hjust = -0.1, size = 2.8
  ) +
  scale_x_continuous(
    labels = label_percent(scale = 1),
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title    = "Cumulative nominal KFE change by NUTS-2 region, 2020-2024",
    subtitle = "26 regions -- Aegean coastal and Ankara region lead",
    x = "Cumulative % change (2020-2024)", y = NULL,
    caption  = "Source: TCMB KFE regional data."
  ) +
  theme(axis.text.y = element_text(size = 8))
```

The Aydin-Denizli-Mugla region (Aegean coast) recorded the highest five-year
cumulative gain at ~2,180%, driven by coastal demand and urban migration from
Istanbul. Mardin-Batman-Sirnak-Siirt lagged at ~1,380%, reflecting lower income
levels. The range across 26 NUTS-2 regions is roughly 1.6x.

### 2.2 Annual KFE change heatmap (2022-2024)

```{r regional-heatmap, fig.height=7}
regional_nuts2 |>
  filter(!is.na(kfe_2022_yoy_pct)) |>
  select(region_name,
         `2022` = kfe_2022_yoy_pct,
         `2023` = kfe_2023_yoy_pct,
         `2024` = kfe_2024_yoy_pct) |>
  pivot_longer(-region_name, names_to = "year_label", values_to = "pct") |>
  filter(!is.na(pct)) |>
  mutate(region_name = fct_reorder(region_name, pct, mean, .na_rm = TRUE)) |>
  ggplot(aes(year_label, region_name, fill = pct)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  geom_text(aes(label = paste0(round(pct, 0), "%")),
            size = 2.5, colour = "white") +
  scale_fill_gradient2(
    low      = pal$secondary,
    mid      = pal$warn,
    high     = pal$accent,
    midpoint = 50,
    name     = "YoY (%)",
    na.value = "grey80"
  ) +
  labs(
    title    = "Annual KFE change by region, 2022-2024 (%)",
    subtitle = "2022 saw nationwide triple-digit gains; cooling is uneven across regions",
    x = NULL, y = NULL,
    caption  = "Source: TCMB KFE. NA = not published by data collection date."
  ) +
  theme(axis.text.y = element_text(size = 8))
```

### 2.3 Median m2 price by region, 2024

```{r regional-prices, fig.height=6.5}
regional_nuts2 |>
  filter(!is.na(median_price_try_m2_2024)) |>
  mutate(region_name = fct_reorder(region_name, median_price_try_m2_2024)) |>
  ggplot(aes(median_price_try_m2_2024, region_name)) +
  geom_col(fill = pal$accent) +
  geom_text(
    aes(label = paste0("TRY ", comma(median_price_try_m2_2024))),
    hjust = -0.08, size = 2.8
  ) +
  scale_x_continuous(
    labels = label_number(big.mark = ","),
    expand = expansion(mult = c(0, 0.22))
  ) +
  labs(
    title    = "Median housing unit price (TRY / m2) by NUTS-2 region, 2024",
    subtitle = "Istanbul commands a 3x premium over Southeast Anatolia regions",
    x = "Median price (TRY / m2)", y = NULL,
    caption  = "Source: TCMB KFE quarterly unit price data."
  ) +
  theme(axis.text.y = element_text(size = 8))
```

---

## 3. Housing Affordability

### 3.1 Months of minimum wage to buy 100 m2

```{r affordability-months, fig.height=4.5}
affordability |>
  ggplot(aes(year, months_100m2)) +
  geom_area(fill = pal$accent, alpha = 0.15) +
  geom_line(colour = pal$accent, linewidth = 1.2) +
  geom_point(colour = pal$accent, size = 2.5) +
  geom_hline(yintercept = 110.4, linetype = "dashed",
             colour = pal$muted, linewidth = 0.6) +
  annotate("text", x = 2010.5, y = 122,
           label = "2010 baseline: 110 months",
           colour = pal$muted, size = 3.5, hjust = 0) +
  scale_x_continuous(breaks = 2010:2024) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "Months of minimum wage required to purchase 100 m2, 2010-2024",
    subtitle = "Affordability worsened 2.5x by 2022; slight recovery in 2024",
    x = NULL, y = "Monthly minimum wages (number of months)",
    caption  = "Source: TCMB KFE unit prices; TURKSTAT minimum wage."
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

In 2010, a minimum-wage worker needed ~110 months (~9 years) of income to buy
100 m2. By 2022 this peaked at ~276 months (~23 years) -- a 2.5x deterioration.
The 2024 reading of ~307 months reflects that prices have continued rising faster
than the minimum wage even after the wage was more than doubled in 2023.

### 3.2 USD-denominated price per m2

```{r usd-prices, fig.height=4}
affordability |>
  ggplot(aes(year, avg_m2_price_usd)) +
  geom_line(colour = pal$secondary, linewidth = 1.2) +
  geom_point(colour = pal$secondary, size = 2.5) +
  geom_area(fill = pal$secondary, alpha = 0.1) +
  scale_x_continuous(breaks = 2010:2024) +
  scale_y_continuous(labels = label_dollar()) +
  labs(
    title    = "Average housing price in USD / m2, 2010-2024",
    subtitle = "Dollar prices collapsed with TRY depreciation in 2018-2022, then partially recovered",
    x = NULL, y = "USD / m2",
    caption  = "Source: TCMB KFE; TCMB USD/TRY average exchange rate."
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### 3.3 Minimum wage vs. m2 price (log scale)

```{r wage-vs-price, fig.height=4}
affordability |>
  select(year,
         `Minimum wage (TRY)` = min_wage_gross_try,
         `Avg. m2 price (TRY)` = avg_m2_price_try) |>
  pivot_longer(-year, names_to = "series", values_to = "value") |>
  ggplot(aes(year, value, colour = series)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  scale_colour_manual(values = c(pal$accent, pal$secondary), name = NULL) +
  scale_x_continuous(breaks = 2010:2024) +
  scale_y_log10(labels = label_number(big.mark = ",")) +
  labs(
    title    = "Minimum wage vs. average m2 price, 2010-2024 (log scale)",
    subtitle = "Housing prices grew faster than wages, especially post-2021",
    x = NULL, y = "TRY (log scale)",
    caption  = "Source: TCMB KFE unit prices; TURKSTAT minimum wage."
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### 3.4 Minimum wage history with mid-year adjustments

```{r minwage-history, fig.height=4}
minwage |>
  ggplot(aes(year, min_wage_gross_try)) +
  geom_col(fill = pal$secondary, alpha = 0.8) +
  geom_point(aes(y = ifelse(!is.na(mid_year_gross_try), mid_year_gross_try, NA)),
             colour = pal$accent, size = 3, na.rm = TRUE) +
  scale_x_continuous(breaks = 2010:2024) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "Minimum wage (gross TRY) per year, 2010-2024",
    subtitle = "Bars = January rate; red dots = mid-year adjustment (when applicable)",
    x = NULL, y = "Gross monthly minimum wage (TRY)",
    caption  = "Source: Turkish Ministry of Labor and Social Security."
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

---

## 4. Exchange Rate and Housing Prices

### 4.1 USD/TRY vs. nominal KFE (quarterly)

```{r fx-kfe, fig.height=4.5}
base_usd <- quarterly_fx$usdtry_avg[1]
base_kfe <- quarterly_fx$kfe_nominal_2010base[1]

quarterly_fx |>
  mutate(
    usdtry_norm = usdtry_avg        / base_usd,
    kfe_norm    = kfe_nominal_2010base / base_kfe
  ) |>
  select(period, `USD/TRY` = usdtry_norm, `KFE nominal` = kfe_norm) |>
  pivot_longer(-period, names_to = "series", values_to = "value") |>
  ggplot(aes(period, value, colour = series, group = series)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  scale_colour_manual(values = c(pal$secondary, pal$accent), name = NULL) +
  scale_y_continuous(labels = label_number(suffix = "x")) +
  labs(
    title    = "USD/TRY vs. KFE -- both normalised to Q1 2015 = 1x",
    subtitle = "Currency depreciation consistently leads or accompanies KFE spikes",
    x = NULL, y = "Index (Q1 2015 = 1x)",
    caption  = "Source: TCMB EVDS (USD/TRY); TCMB KFE."
  ) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 7))
```

### 4.2 YoY change correlation: FX vs. KFE

```{r fx-correlation, fig.height=4}
quarterly_fx |>
  filter(!is.na(kfe_yoy_pct)) |>
  mutate(usdtry_yoy = (usdtry_avg / lag(usdtry_avg, 4) - 1) * 100) |>
  filter(!is.na(usdtry_yoy)) |>
  ggplot(aes(usdtry_yoy, kfe_yoy_pct)) +
  geom_point(aes(colour = factor(year)), size = 3) +
  geom_smooth(method = "lm", se = TRUE,
              colour = pal$accent, fill = pal$accent,
              alpha = 0.1, linewidth = 0.8) +
  scale_colour_viridis_d(name = "Year", option = "D") +
  scale_x_continuous(labels = label_percent(scale = 1)) +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  labs(
    title    = "USD/TRY annual change vs. KFE annual change, quarterly 2016-2024",
    subtitle = "Strong positive correlation -- currency depreciation feeds into housing prices",
    x = "USD/TRY YoY change (%)", y = "KFE nominal YoY change (%)",
    caption  = "Source: TCMB EVDS; TCMB KFE. Each point = one quarter."
  )
```

### 4.3 Real vs. nominal KFE change over time

```{r real-premium, fig.height=4}
quarterly_fx |>
  filter(!is.na(kfe_yoy_pct)) |>
  select(period, year,
         `Nominal KFE YoY` = kfe_yoy_pct,
         `CPI YoY`         = cpi_yoy_pct,
         `Real KFE YoY`    = kfe_real_yoy_pct) |>
  pivot_longer(-c(period, year), names_to = "series", values_to = "pct") |>
  ggplot(aes(period, pct, colour = series, group = series)) +
  geom_line(linewidth = 0.9) +
  geom_hline(yintercept = 0, linetype = "dashed",
             colour = pal$muted, linewidth = 0.4) +
  scale_colour_manual(
    values = c(pal$muted, pal$accent, pal$secondary),
    name   = NULL
  ) +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  labs(
    title    = "Nominal KFE, CPI, and real KFE change, quarterly 2016-2024",
    subtitle = "Real KFE has been negative in most quarters since late 2022",
    x = NULL, y = "YoY change (%)",
    caption  = "Source: TCMB KFE; TURKSTAT CPI."
  ) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 7))
```

---

## 5. Supply Side: Sales and Building Permits

### 5.1 Housing sales by type

```{r sales-trend, fig.height=4.5}
housing_sales |>
  select(year,
         `Total sales`  = sales_total,
         `First-hand`   = sales_first_hand,
         `Second-hand`  = sales_second_hand) |>
  filter(year >= 2015) |>
  pivot_longer(-year, names_to = "type", values_to = "count") |>
  filter(!is.na(count)) |>
  ggplot(aes(year, count, colour = type)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_colour_manual(
    values = c(pal$accent, pal$secondary, pal$warn),
    name   = NULL
  ) +
  scale_x_continuous(breaks = 2015:2024) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "Housing sales by type, 2015-2024",
    subtitle = "Total sales peaked in 2020 (credit stimulus); mortgage share collapsed post-2022",
    x = NULL, y = "Number of units sold",
    caption  = "Source: TUIK Insaat ve Konut Istatistikleri."
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Sales peaked in 2020 at 1.5 million units, boosted by the government mortgage
campaign. Mortgage sales collapsed from 601,466 in 2020 to just 158,486 in 2024
as interest rates rose steeply, with mortgage share falling from 40% to 12%.

### 5.2 Foreign buyer sales

```{r foreign-sales, fig.height=3.5}
housing_sales |>
  filter(!is.na(sales_foreign)) |>
  ggplot(aes(year, sales_foreign)) +
  geom_col(fill = pal$secondary) +
  geom_text(aes(label = comma(sales_foreign)), vjust = -0.4, size = 3) +
  scale_x_continuous(breaks = 2010:2024) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.1))) +
  labs(
    title    = "Foreign buyer housing sales, 2010-2024",
    subtitle = "Peaked at 67,490 in 2022 after citizenship-by-investment boom; declined sharply after",
    x = NULL, y = "Units sold to foreign buyers",
    caption  = "Source: TUIK."
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### 5.3 Building permits: the supply pipeline

```{r permits-trend, fig.height=4}
permits |>
  select(year,
         `Permits (units)` = permits_residential_units,
         `Occupancy (units)` = occupancy_permits_units) |>
  pivot_longer(-year, names_to = "series", values_to = "value") |>
  ggplot(aes(year, value, colour = series)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.2) +
  scale_colour_manual(values = c(pal$secondary, pal$accent), name = NULL) +
  scale_x_continuous(breaks = 2010:2024) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "Residential building permits and occupancy certificates, 2010-2024",
    subtitle = "Permits halved since 2015 peak -- structural undersupply through 2026-2027",
    x = NULL, y = "Number of units",
    caption  = "Source: TUIK Insaat ve Konut Istatistikleri."
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Building permits fell from 1.22 million units in 2015 to 549,000 in 2024, a
55% decline. Given the 2-4 year lag between permit issuance and delivery, this
signals structural supply shortfall through at least 2026-2027.

### 5.4 Interactive full sales and permits table

```{r sales-permits-dt}
annual_macro |>
  select(year, housing_sales_total, housing_sales_mortgage,
         housing_sales_foreign, building_permits_residential_units) |>
  mutate(
    housing_sales_total                = comma(housing_sales_total),
    housing_sales_mortgage             = ifelse(is.na(housing_sales_mortgage), "--",
                                                comma(housing_sales_mortgage)),
    housing_sales_foreign              = comma(housing_sales_foreign),
    building_permits_residential_units = comma(building_permits_residential_units)
  ) |>
  datatable(
    rownames = FALSE,
    filter   = "top",
    options  = list(pageLength = 15, dom = "ftp"),
    caption  = "Annual housing sales and supply indicators, 2010-2024"
  )
```

---

## Conclusions

Four structural findings emerge from this analysis:

1. **The Turkish housing boom is largely nominal, not real.** KFE rose ~48x
   nominally but the real index peaked in 2022 and has since declined.
   Buyers in 2021-2023 effectively paid an inflation premium.

2. **Affordability has collapsed to crisis levels.** The minimum-wage-months
   metric worsened 2.5x by 2022 and continues to deteriorate, reflecting a
   decade of supply under-investment plus income growth unable to keep pace
   with price inflation or currency depreciation.

3. **Exchange rate risk is the dominant driver.** USD/TRY depreciation
   episodes directly precede or coincide with the sharpest nominal KFE
   increases. Turkish housing has acted as a partial inflation hedge in lira
   terms, drawing domestic savings away from financial assets.

4. **The supply pipeline signals prolonged undersupply.** The 55% fall in
   building permits since 2015, combined with post-earthquake relocation
   demand and collapsed mortgage activity, means structural supply constraints
   will persist even if nominal price growth slows further.
