---
title: "Analysis"
subtitle: "Nominal Surge, Real Stagnation, and the Affordability Crisis 2010-2024"
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(scales)
library(knitr)
library(tidyr)
library(forcats)
library(stringr)
library(DT)

load("data/housing_market_tr.RData")

pal <- list(
  primary   = "#1a1814",
  secondary = "#2c3e50",
  accent    = "#c0392b",
  good      = "#27ae60",
  warn      = "#e67e22",
  muted     = "#95a5a6"
)

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title       = element_text(face = "bold", size = 13, hjust = 0),
      plot.subtitle    = element_text(colour = "#6b6560", size = 10.5, hjust = 0),
      plot.caption     = element_text(colour = "#95a5a6", size = 8, hjust = 0),
      plot.background  = element_rect(fill = "#f8f5f0", colour = NA),
      panel.background = element_rect(fill = "#f8f5f0", colour = NA),
      panel.grid.major = element_line(colour = "#e0dbd3"),
      panel.grid.minor = element_blank(),
      legend.position  = "top",
      legend.text      = element_text(size = 9),
      axis.text        = element_text(colour = "#6b6560", size = 9),
      axis.title       = element_text(colour = "#1a1814", size = 10)
    )
)
```

::: {.callout-note icon="false"}
### Key Takeaways

- **Nominal surge, real stagnation.** KFE rose ~4,800% nominally (2010-2024);
  in real terms the index peaked at 153.7 in 2022 and fell to 118.4 by 2024.
- **Affordability collapsed.** Months of minimum wage to buy 100 m2 rose from
  103 in 2010 to 330 in 2023 -- a 3x worsening in 13 years.
- **Exchange rate pass-through is strong.** Quarterly correlations between
  USD/TRY and KFE changes exceed 0.70.
- **Istanbul and Aegean lead, Southeast lags.** Five-year cumulative nominal
  gains range from ~2,070% (Aegean) to ~1,420% (Southeast Anatolia).
- **Supply is contracting.** Building permits fell from 1.22 million units in
  2015 to 549,000 in 2024 -- a 55% drop.
:::

---

## 1. Nominal vs. Real Price Evolution

### 1.1 The two stories of Turkish housing prices

```{r nominal-vs-real, fig.height=4.5}
kfe_annual |>
  select(year, `Nominal KFE` = kfe_nominal, `Real KFE` = kfe_real) |>
  pivot_longer(-year, names_to = "series", values_to = "value") |>
  ggplot(aes(year, value, colour = series, linetype = series)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_colour_manual(values = c(pal$accent, pal$secondary), name = NULL) +
  scale_linetype_manual(values = c("solid", "dashed"), name = NULL) +
  scale_x_continuous(breaks = 2010:2024) +
  scale_y_continuous(labels = comma) +
  labs(
    title   = "Nominal vs. real KFE, 2010-2024 (2010 = 100)",
    subtitle = "Nominal growth is dramatic; real gains are modest and recently reversed",
    x = NULL, y = "Index (2010 = 100)",
    caption = "Source: TCMB KFE; deflated using TURKSTAT CPI."
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The two lines tell a strikingly different story. While the nominal index rose
~48x from 2010 to 2024, the real index peaked at 153.7 in 2022 -- just 54%
above its 2010 level -- before declining. By 2024, real house prices were lower
than in 2022, meaning buyers who purchased at the peak paid a substantial
inflation premium with no real asset gain.

### 1.2 Annual change rates

```{r yoy-change, fig.height=4}
kfe_annual |>
  filter(!is.na(kfe_nominal_yoy)) |>
  select(year,
         `KFE nominal YoY %` = kfe_nominal_yoy,
         `CPI YoY %`         = cpi_yoy,
         `KFE real YoY %`    = kfe_real_yoy) |>
  pivot_longer(-year, names_to = "series", values_to = "pct") |>
  ggplot(aes(year, pct, fill = series)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_hline(yintercept = 0, colour = pal$primary, linewidth = 0.4) +
  scale_fill_manual(
    values = c(pal$accent, pal$muted, pal$secondary),
    name   = NULL
  ) +
  scale_x_continuous(breaks = 2011:2024) +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  labs(
    title   = "Annual price change: KFE vs. CPI, 2011-2024",
    subtitle = "Real KFE turned negative in 2018 and again in 2023-2024",
    x = NULL, y = "Year-on-year change (%)",
    caption = "Source: TCMB KFE, TURKSTAT CPI."
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The real KFE went negative in two distinct episodes: 2018-2019 after the
currency crisis, and again in 2023-2024 as inflation ran ahead of price
growth. In 2022 the nominal KFE surged ~183% YoY, yet CPI at ~85% limited
real gains to ~53%.

### 1.3 Cumulative growth: housing vs. inflation

```{r cumulative, fig.height=4}
kfe_annual |>
  mutate(
    cpi_cumul = cpi_index / 100,
    kfe_cumul = kfe_nominal / 100
  ) |>
  select(year, `Nominal KFE` = kfe_cumul, `CPI` = cpi_cumul) |>
  pivot_longer(-year) |>
  ggplot(aes(year, value, colour = name)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_colour_manual(values = c(pal$muted, pal$accent), name = NULL) +
  scale_x_continuous(breaks = 2010:2024) +
  scale_y_continuous(labels = label_number(suffix = "x")) +
  labs(
    title   = "Cumulative multiplier since 2010: KFE vs. general inflation",
    subtitle = "In 2022-2023, housing outpaced CPI; by 2024 the gap is closing",
    x = NULL, y = "Cumulative multiplier (2010 = 1x)",
    caption = "Source: TCMB KFE; TURKSTAT CPI."
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

---

## 2. Regional Analysis

### 2.1 Five-year cumulative nominal change (2020-2024)

```{r regional-bar, fig.height=5}
regional_kfe |>
  mutate(region_name = fct_reorder(region_name, cumulative_2020_2024_pct)) |>
  ggplot(aes(cumulative_2020_2024_pct, region_name)) +
  geom_col(fill = pal$secondary) +
  geom_text(
    aes(label = paste0(round(cumulative_2020_2024_pct, 0), "%")),
    hjust = -0.1, size = 3.2
  ) +
  scale_x_continuous(
    labels = label_percent(scale = 1),
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title   = "Cumulative nominal KFE change by NUTS-1 region, 2020-2024",
    subtitle = "Five-year compounded gain -- Aegean and West Anatolia lead",
    x = "Cumulative % change", y = NULL,
    caption = "Source: TCMB KFE regional data."
  )
```

The Aegean region (Izmir, Mugla, Aydin) recorded the highest five-year
cumulative gain at ~2,070%, driven by strong coastal demand and urban
migration. Southeast Anatolia lagged at ~1,420%, reflecting lower income
levels and weaker investor demand.

### 2.2 Annual regional heatmap (2022-2024)

```{r regional-heatmap, fig.height=5}
regional_kfe |>
  select(region_name,
         `2022` = kfe_2022_pct,
         `2023` = kfe_2023_pct,
         `2024` = kfe_2024_pct) |>
  pivot_longer(-region_name, names_to = "year", values_to = "pct") |>
  mutate(region_name = fct_reorder(region_name, pct, mean)) |>
  ggplot(aes(year, region_name, fill = pct)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = paste0(round(pct, 0), "%")),
            size = 3.2, colour = "white") +
  scale_fill_gradient2(
    low      = pal$secondary,
    mid      = pal$warn,
    high     = pal$accent,
    midpoint = 50,
    name     = "YoY (%)"
  ) +
  labs(
    title   = "Annual KFE change by region, 2022-2024 (%)",
    subtitle = "2022 saw nationwide triple-digit gains; cooling is visible but uneven",
    x = NULL, y = NULL,
    caption = "Source: TCMB KFE."
  )
```

### 2.3 Median m2 price by region, 2024

```{r regional-prices, fig.height=5}
regional_kfe |>
  mutate(region_name = fct_reorder(region_name, median_price_2024_try_m2)) |>
  ggplot(aes(median_price_2024_try_m2, region_name)) +
  geom_col(fill = pal$accent) +
  geom_text(
    aes(label = paste0("TRY ", comma(median_price_2024_try_m2))),
    hjust = -0.1, size = 3.2
  ) +
  scale_x_continuous(
    labels = label_number(big.mark = ","),
    expand = expansion(mult = c(0, 0.2))
  ) +
  labs(
    title   = "Median housing unit price (TRY / m2) by NUTS-1 region, 2024",
    subtitle = "Istanbul commands a 3x premium over Southeast Anatolia",
    x = "Median price (TRY / m2)", y = NULL,
    caption = "Source: TCMB KFE quarterly unit price data."
  )
```

Istanbul's median of TRY 95,420/m2 is 3.4x Southeast Anatolia's TRY 28,150/m2,
reflecting deep structural differences in income, urbanisation density, and
investment demand.

---

## 3. Housing Affordability

### 3.1 Months of minimum wage to buy 100 m2

```{r affordability-months, fig.height=4.5}
affordability |>
  ggplot(aes(year, affordability_months)) +
  geom_area(fill = pal$accent, alpha = 0.15) +
  geom_line(colour = pal$accent, linewidth = 1.2) +
  geom_point(colour = pal$accent, size = 2.5) +
  geom_hline(yintercept = 103.2, linetype = "dashed",
             colour = pal$muted, linewidth = 0.6) +
  annotate("text", x = 2010.5, y = 115,
           label = "2010 baseline: 103 months",
           colour = pal$muted, size = 3.5, hjust = 0) +
  scale_x_continuous(breaks = 2010:2024) +
  scale_y_continuous(labels = comma) +
  labs(
    title   = "Months of minimum wage required to purchase 100 m2, 2010-2024",
    subtitle = "Affordability worsened 3x in 13 years -- a structural housing crisis",
    x = NULL, y = "Monthly minimum wages (number of months)",
    caption = "Source: TCMB KFE unit prices; TURKSTAT minimum wage series."
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

In 2010, a Turkish worker on minimum wage needed ~103 months (~8.6 years) of
income to buy a 100 m2 flat. By 2023 this had risen to ~330 months (~27.5
years) -- a 3.2x deterioration. The slight improvement in 2024 reflects
minimum wage increases outpacing nominal KFE growth that year, but
affordability remains structurally impaired.

### 3.2 USD-denominated prices

```{r usd-prices, fig.height=4}
affordability |>
  ggplot(aes(year, m2_usd)) +
  geom_line(colour = pal$secondary, linewidth = 1.2) +
  geom_point(colour = pal$secondary, size = 2.5) +
  geom_area(fill = pal$secondary, alpha = 0.1) +
  scale_x_continuous(breaks = 2010:2024) +
  scale_y_continuous(labels = label_dollar()) +
  labs(
    title   = "Average housing price in USD / m2, 2010-2024",
    subtitle = "Dollar prices peaked in 2024 after collapsing with TRY depreciation in 2019-2022",
    x = NULL, y = "USD / m2",
    caption = "Source: TCMB KFE; TCMB USD/TRY average exchange rate."
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### 3.3 Minimum wage vs. average m2 price

```{r wage-vs-price, fig.height=4}
affordability |>
  select(year,
         `Minimum wage (TRY)` = min_wage_monthly_try,
         `Avg. m2 price (TRY)` = avg_m2_price_try) |>
  pivot_longer(-year, names_to = "series", values_to = "value") |>
  ggplot(aes(year, value, colour = series)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  scale_colour_manual(values = c(pal$accent, pal$secondary), name = NULL) +
  scale_x_continuous(breaks = 2010:2024) +
  scale_y_log10(labels = label_number(big.mark = ",")) +
  labs(
    title   = "Minimum wage vs. average m2 price, 2010-2024 (log scale)",
    subtitle = "Housing prices grew faster than wages, especially post-2021",
    x = NULL, y = "TRY (log scale)",
    caption = "Source: TCMB KFE unit prices; TURKSTAT minimum wage."
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

---

## 4. Exchange Rate and Housing Price Relationship

### 4.1 USD/TRY vs. nominal KFE (quarterly)

```{r fx-kfe, fig.height=4.5}
base_usd <- exchange_housing$usdtry[1]
base_kfe <- exchange_housing$kfe_nominal[1]

exchange_housing |>
  mutate(
    usdtry_norm = usdtry / base_usd,
    kfe_norm    = kfe_nominal / base_kfe
  ) |>
  select(period, `USD/TRY` = usdtry_norm, `KFE nominal` = kfe_norm) |>
  pivot_longer(-period, names_to = "series", values_to = "value") |>
  ggplot(aes(period, value, colour = series, group = series)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  scale_colour_manual(values = c(pal$secondary, pal$accent), name = NULL) +
  scale_y_continuous(labels = label_number(suffix = "x")) +
  labs(
    title   = "USD/TRY vs. KFE -- both normalised to Q1 2018 = 1x",
    subtitle = "Currency depreciation consistently leads or accompanies KFE spikes",
    x = NULL, y = "Index (Q1 2018 = 1x)",
    caption = "Source: TCMB EVDS (USD/TRY); TCMB KFE."
  ) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 7))
```

### 4.2 YoY change correlation

```{r fx-correlation, fig.height=4}
exchange_housing |>
  filter(!is.na(kfe_yoy_pct)) |>
  mutate(usdtry_yoy = (usdtry / lag(usdtry, 4) - 1) * 100) |>
  filter(!is.na(usdtry_yoy)) |>
  ggplot(aes(usdtry_yoy, kfe_yoy_pct)) +
  geom_point(aes(colour = factor(year)), size = 3) +
  geom_smooth(method = "lm", se = TRUE,
              colour = pal$accent, fill = pal$accent, alpha = 0.1,
              linewidth = 0.8) +
  scale_colour_viridis_d(name = "Year", option = "D") +
  scale_x_continuous(labels = label_percent(scale = 1)) +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  labs(
    title   = "USD/TRY annual change vs. KFE annual change, quarterly 2019-2024",
    subtitle = "Strong positive correlation: currency depreciation feeds into housing prices",
    x = "USD/TRY YoY change (%)", y = "KFE nominal YoY change (%)",
    caption = "Source: TCMB EVDS; TCMB KFE. Each point = one quarter."
  )
```

### 4.3 Real vs. nominal KFE growth

```{r real-premium, fig.height=4}
exchange_housing |>
  filter(!is.na(kfe_yoy_pct)) |>
  select(period, year,
         `Nominal KFE YoY` = kfe_yoy_pct,
         `CPI YoY`         = cpi_yoy_pct,
         `Real KFE YoY`    = kfe_real_yoy_pct) |>
  pivot_longer(-c(period, year), names_to = "series", values_to = "pct") |>
  ggplot(aes(period, pct, colour = series, group = series)) +
  geom_line(linewidth = 0.9) +
  geom_hline(yintercept = 0, linetype = "dashed",
             colour = pal$muted, linewidth = 0.4) +
  scale_colour_manual(
    values = c(pal$muted, pal$accent, pal$secondary),
    name   = NULL
  ) +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  labs(
    title   = "Nominal KFE, CPI, and real KFE change, quarterly 2019-2024",
    subtitle = "Real KFE has been negative in most quarters since Q3 2022",
    x = NULL, y = "YoY change (%)",
    caption = "Source: TCMB KFE; TURKSTAT CPI."
  ) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 7))
```

---

## 5. Supply Side: Sales and Building Permits

### 5.1 Housing sales by type

```{r sales-trend, fig.height=4.5}
sales_permits |>
  select(year,
         `Total sales`  = sales_total,
         `First-hand`   = sales_first_hand,
         `Second-hand`  = sales_second_hand) |>
  filter(year >= 2015) |>
  pivot_longer(-year, names_to = "type", values_to = "count") |>
  filter(!is.na(count)) |>
  ggplot(aes(year, count, colour = type)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_colour_manual(
    values = c(pal$accent, pal$secondary, pal$warn),
    name   = NULL
  ) +
  scale_x_continuous(breaks = 2015:2024) +
  scale_y_continuous(labels = comma) +
  labs(
    title   = "Housing sales by type, 2015-2024",
    subtitle = "Total sales peaked in 2020 (credit stimulus) and have since declined",
    x = NULL, y = "Number of units sold",
    caption = "Source: TUIK Insaat ve Konut Istatistikleri."
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Sales peaked in 2020 at 1.5 million units, boosted by the government's
subsidised mortgage campaign. Since then, rising prices and tightening credit
have driven sales down to 1.07 million in 2024 -- the lowest since 2010.

### 5.2 Building permits: the supply pipeline

```{r permits-trend, fig.height=4}
sales_permits |>
  select(year,
         `Building permits (units)` = permits_residential_units,
         `Total sales`              = sales_total) |>
  pivot_longer(-year, names_to = "series", values_to = "value") |>
  ggplot(aes(year, value, colour = series)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.2) +
  scale_colour_manual(values = c(pal$accent, pal$secondary), name = NULL) +
  scale_x_continuous(breaks = 2010:2024) +
  scale_y_continuous(labels = comma) +
  labs(
    title   = "Building permits vs. housing sales, 2010-2024",
    subtitle = "Supply pipeline halved from 2015 peak -- structural undersupply ahead",
    x = NULL, y = "Number of units",
    caption = "Source: TUIK Insaat ve Konut Istatistikleri."
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Building permits fell from 1.22 million units in 2015 to 549,000 in 2024, a
55% decline. Given the 2-4 year lag between permit issuance and delivery,
this signals structural supply shortfall through at least 2026-2027.

### 5.3 Interactive data table

```{r sales-dt}
sales_permits |>
  mutate(
    sales_total               = comma(sales_total),
    permits_residential_units = comma(permits_residential_units),
    second_hand_share         = ifelse(is.na(second_hand_share), "--",
                                       paste0(round(second_hand_share, 1), "%")),
    mortgage_share            = ifelse(is.na(mortgage_share), "--",
                                       paste0(round(mortgage_share, 1), "%"))
  ) |>
  select(year, sales_total, permits_residential_units,
         second_hand_share, mortgage_share, sales_foreign) |>
  datatable(
    rownames = FALSE,
    filter   = "top",
    options  = list(pageLength = 15, dom = "ftp"),
    caption  = "Annual housing market supply-demand indicators, 2010-2024"
  )
```

---

## Conclusions

Four structural findings emerge:

1. **The Turkish housing boom is largely nominal, not real.** KFE rose ~48x
   nominally but only ~18% in real terms since 2010, and that real gain has
   partially reversed since 2022.

2. **Affordability has collapsed to crisis levels.** The 3x worsening in the
   months-of-minimum-wage metric is structural -- it reflects a decade of
   supply under-investment coinciding with income growth that could not keep
   pace with price inflation or currency depreciation.

3. **Exchange rate risk is the dominant driver.** The pass-through from USD/TRY
   depreciation to nominal housing prices is rapid and strong. Turkish housing
   has effectively acted as a partial inflation hedge in lira terms.

4. **The supply pipeline threatens prolonged undersupply.** The 55% fall in
   building permits since 2015, combined with post-earthquake relocation demand,
   means structural supply constraints will persist even if nominal price growth
   slows further.
