---
title: "Analysis"
subtitle: "Nominal Surge, Real Stagnation, and the Affordability Crisis — 2010–2024"
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(scales)
library(knitr)
library(tidyr)
library(forcats)
library(stringr)
library(DT)

load("data/housing_market_tr.RData")

pal <- list(
  primary   = "#1a1814",
  secondary = "#2c3e50",
  accent    = "#c0392b",
  good      = "#27ae60",
  warn      = "#e67e22",
  muted     = "#95a5a6"
)

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title       = element_text(face = "bold", size = 13, hjust = 0),
      plot.subtitle    = element_text(colour = "#6b6560", size = 10.5, hjust = 0),
      plot.caption     = element_text(colour = "#95a5a6", size = 8, hjust = 0),
      plot.background  = element_rect(fill = "#f8f5f0", colour = NA),
      panel.background = element_rect(fill = "#f8f5f0", colour = NA),
      panel.grid.major = element_line(colour = "#e0dbd3"),
      panel.grid.minor = element_blank(),
      legend.position  = "top",
      legend.text      = element_text(size = 9),
      axis.text        = element_text(colour = "#6b6560", size = 9),
      axis.title       = element_text(colour = "#1a1814", size = 10)
    )
)
```

::: {.callout-note icon="false"}
### Key Takeaways

- **Nominal surge, real stagnation.** KFE rose ~4,800% nominally (2010–2024); in real
  terms the index peaked at 153.7 in 2022 and fell back to 118.4 by 2024, suggesting
  the surge was largely inflation-driven.
- **Affordability collapsed.** The months-of-minimum-wage needed to buy 100 m² rose
  from ~103 in 2010 to ~330 in 2023 — a 3× worsening in a decade.
- **Exchange rate shocks transmit into housing prices.** Quarterly correlations between
  USD/TRY change and KFE change exceed 0.70 — currency depreciation episodes directly
  precede or accompany the sharpest price spikes.
- **Istanbul and Aegean lead, Southeast lags.** Five-year cumulative nominal gains
  range from ~2,000% (Aegean) to ~1,400% (Southeast Anatolia).
- **Supply is in structural decline.** Building permits halved from 2015 peak to 2024,
  compressing supply just as demand surged post-earthquake.
:::

---

## 1. Nominal vs. Real Price Evolution

### 1.1 The two stories of Turkish housing prices

```{r nominal-vs-real, fig.height=4.5}
kfe_annual |>
  select(year, `Nominal KFE` = kfe_nominal, `Real KFE` = kfe_real) |>
  pivot_longer(-year, names_to = "series", values_to = "value") |>
  ggplot(aes(year, value, colour = series, linetype = series)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_colour_manual(values = c(pal$accent, pal$secondary), name = NULL) +
  scale_linetype_manual(values = c("solid", "dashed"), name = NULL) +
  scale_x_continuous(breaks = 2010:2024) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "Nominal vs. real KFE, 2010–2024 (2010 = 100)",
    subtitle = "Nominal growth is dramatic; real gains are modest and recently reversed",
    x = NULL, y = "Index (2010 = 100)",
    caption  = "Source: TCMB KFE; deflated using TURKSTAT CPI."
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The two lines tell a strikingly different story. While the **nominal index** rose
~48× from 2010 to 2024, the **real index** peaked at 153.7 in 2022 — just 54% above
its 2010 level — before declining. By 2024, real house prices were lower than in 2022,
meaning buyers who purchased at the peak paid a substantial inflation premium with no
real asset gain.

### 1.2 Annual change rates

```{r yoy-change, fig.height=4}
kfe_annual |>
  filter(!is.na(kfe_nominal_yoy)) |>
  select(year, `KFE nominal YoY %` = kfe_nominal_yoy,
               `CPI YoY %`         = cpi_yoy,
               `KFE real YoY %`    = kfe_real_yoy) |>
  pivot_longer(-year, names_to = "series", values_to = "pct") |>
  ggplot(aes(year, pct, fill = series)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_hline(yintercept = 0, colour = pal$primary, linewidth = 0.4) +
  scale_fill_manual(
    values = c(pal$accent, pal$muted, pal$secondary),
    name   = NULL
  ) +
  scale_x_continuous(breaks = 2011:2024) +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  labs(
    title    = "Annual price change: KFE vs. CPI, 2011–2024",
    subtitle = "Real KFE turned negative in 2018 and again in 2023–2024",
    x = NULL, y = "Year-on-year change (%)",
    caption  = "Source: TCMB KFE, TURKSTAT CPI."
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The **real KFE went negative** in two distinct episodes: 2018–2019 (after the currency
crisis) and again in 2023–2024 (as inflation ran ahead of price growth). In 2022 the
nominal KFE surged ~183% YoY — yet even then, the CPI surge of ~85% limited real gains
to ~53%. The 2024 real decline of ~8% suggests the boom is ending in real terms.

### 1.3 Cumulative growth: housing vs. inflation

```{r cumulative, fig.height=4}
kfe_annual |>
  mutate(
    cpi_cumul = cpi_index / 100,   # 2010=1
    kfe_cumul = kfe_nominal / 100  # 2010=1
  ) |>
  select(year, `Nominal KFE` = kfe_cumul, `CPI` = cpi_cumul) |>
  pivot_longer(-year) |>
  ggplot(aes(year, value, colour = name)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_colour_manual(values = c(pal$muted, pal$accent), name = NULL) +
  scale_x_continuous(breaks = 2010:2024) +
  scale_y_continuous(labels = label_number(suffix = "×")) +
  labs(
    title    = "Cumulative multiplier since 2010 — KFE vs. general inflation",
    subtitle = "In 2022–2023, housing outpaced CPI; by 2024 the gap is closing",
    x = NULL, y = "Cumulative multiplier (2010 = 1×)",
    caption  = "Source: TCMB KFE; TURKSTAT CPI."
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

---

## 2. Regional Analysis

### 2.1 Five-year cumulative nominal change (2020–2024)

```{r regional-bar, fig.height=5}
regional_kfe |>
  mutate(region_name = fct_reorder(region_name, cumulative_2020_2024_pct)) |>
  ggplot(aes(cumulative_2020_2024_pct, region_name)) +
  geom_col(fill = pal$secondary) +
  geom_text(
    aes(label = paste0(round(cumulative_2020_2024_pct, 0), "%")),
    hjust = -0.1, size = 3.2
  ) +
  scale_x_continuous(
    labels = label_percent(scale = 1),
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title    = "Cumulative nominal KFE change by NUTS-1 region, 2020–2024",
    subtitle = "Five-year compounded gain — Aegean and West Anatolia lead",
    x = "Cumulative % change", y = NULL,
    caption  = "Source: TCMB KFE regional data."
  )
```

The **Aegean region** (İzmir, Muğla, Aydın) recorded the highest five-year cumulative
gain at ~2,070%, driven by strong coastal demand and urban migration. **Southeast
Anatolia** lagged at ~1,420%, reflecting lower income levels and weaker investor
demand. The range is roughly **1.5× between the fastest and slowest region**.

### 2.2 Annual regional comparison (2022–2024)

```{r regional-heatmap, fig.height=5}
regional_kfe |>
  select(region_name, `2022` = kfe_2022_pct, `2023` = kfe_2023_pct, `2024` = kfe_2024_pct) |>
  pivot_longer(-region_name, names_to = "year", values_to = "pct") |>
  mutate(region_name = fct_reorder(region_name, pct, mean)) |>
  ggplot(aes(year, region_name, fill = pct)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = paste0(round(pct, 0), "%")), size = 3.2, colour = "white") +
  scale_fill_gradient2(
    low     = pal$secondary,
    mid     = pal$warn,
    high    = pal$accent,
    midpoint = 50,
    name    = "YoY change (%)"
  ) +
  labs(
    title    = "Annual KFE change by region, 2022–2024 (%)",
    subtitle = "2022 saw nationwide triple-digit gains; cooling is visible but uneven",
    x = NULL, y = NULL,
    caption  = "Source: TCMB KFE."
  )
```

### 2.3 Median m² price by region, 2024

```{r regional-prices, fig.height=5}
regional_kfe |>
  mutate(region_name = fct_reorder(region_name, median_price_2024_try_m2)) |>
  ggplot(aes(median_price_2024_try_m2, region_name)) +
  geom_col(fill = pal$accent) +
  geom_text(
    aes(label = paste0("₺", comma(median_price_2024_try_m2))),
    hjust = -0.1, size = 3.2
  ) +
  scale_x_continuous(
    labels = label_number(prefix = "₺", big.mark = ","),
    expand = expansion(mult = c(0, 0.2))
  ) +
  labs(
    title    = "Median housing unit price (TRY / m²) by NUTS-1 region, 2024",
    subtitle = "Istanbul commands a 3× premium over Southeast Anatolia",
    x = "Median price (TRY / m²)", y = NULL,
    caption  = "Source: TCMB KFE quarterly unit price data."
  )
```

Istanbul's median of **₺95,420/m²** is **3.4×** Southeast Anatolia's ₺28,150/m² —
reflecting deep structural differences in income, urbanisation density, and investment
demand.

---

## 3. Housing Affordability

### 3.1 Months of minimum wage to buy 100 m²

```{r affordability-months, fig.height=4.5}
affordability |>
  ggplot(aes(year, affordability_months)) +
  geom_area(fill = pal$accent, alpha = 0.15) +
  geom_line(colour = pal$accent, linewidth = 1.2) +
  geom_point(colour = pal$accent, size = 2.5) +
  geom_hline(yintercept = 103.2, linetype = "dashed",
             colour = pal$muted, linewidth = 0.6) +
  annotate("text", x = 2010.5, y = 110, label = "2010 baseline: 103 months",
           colour = pal$muted, size = 3.5, hjust = 0) +
  scale_x_continuous(breaks = 2010:2024) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "Months of minimum wage required to purchase 100 m², 2010–2024",
    subtitle = "Affordability worsened 3× in 13 years — a structural housing crisis",
    x = NULL, y = "Monthly minimum wages (# of months)",
    caption  = "Source: TCMB KFE unit prices; TURKSTAT minimum wage series."
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

In 2010, a Turkish worker on minimum wage needed ~**103 months** (~8.6 years) of income
to buy a 100 m² flat. By 2023 this had risen to **~330 months** (~27.5 years) — a
3.2× deterioration. The slight improvement in 2024 reflects minimum wage increases
outpacing nominal KFE growth that year, but affordability remains structurally impaired.

### 3.2 USD-denominated prices — the dollarisation lens

```{r usd-prices, fig.height=4}
affordability |>
  ggplot(aes(year, m2_usd)) +
  geom_line(colour = pal$secondary, linewidth = 1.2) +
  geom_point(colour = pal$secondary, size = 2.5) +
  geom_area(fill = pal$secondary, alpha = 0.1) +
  scale_x_continuous(breaks = 2010:2024) +
  scale_y_continuous(labels = label_dollar()) +
  labs(
    title    = "Average housing price in USD / m², 2010–2024",
    subtitle = "Dollar prices rose then collapsed with TRY depreciation — real foreign-currency affordability improved",
    x = NULL, y = "USD / m²",
    caption  = "Source: TCMB KFE; TCMB USD/TRY average exchange rate."
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Viewed in USD, the picture inverts: **dollar-denominated prices peaked at ~$918/m² in
2022** then fell as TRY depreciated faster than nominal prices rose. By 2024 the USD
price had partially recovered to ~$1,587. This bifurcation — lira prices rising while
dollar prices fluctuate — captures the inflation illusion at the heart of the Turkish
housing boom.

### 3.3 Min. wage vs. avg. m² price over time

```{r wage-vs-price, fig.height=4}
affordability |>
  select(year,
         `Minimum wage (TRY)` = min_wage_monthly_try,
         `Avg. m² price (TRY)` = avg_m2_price_try) |>
  pivot_longer(-year, names_to = "series", values_to = "value") |>
  ggplot(aes(year, value, colour = series)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  scale_colour_manual(values = c(pal$accent, pal$secondary), name = NULL) +
  scale_x_continuous(breaks = 2010:2024) +
  scale_y_log10(labels = label_number(prefix = "₺", big.mark = ",")) +
  labs(
    title    = "Minimum wage vs. average m² price, 2010–2024 (log scale)",
    subtitle = "Housing prices grew faster than wages — especially post-2021",
    x = NULL, y = "TRY (log scale)",
    caption  = "Source: TCMB KFE unit prices; TURKSTAT minimum wage."
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

---

## 4. Exchange Rate & Housing Price Relationship

### 4.1 USD/TRY vs. nominal KFE — quarterly

```{r fx-kfe, fig.height=4.5}
exchange_housing |>
  select(period, year, quarter,
         `USD/TRY` = usdtry,
         `KFE nominal` = kfe_nominal) |>
  pivot_longer(c(`USD/TRY`, `KFE nominal`), names_to = "series", values_to = "value") |>
  mutate(
    value_norm = ifelse(series == "USD/TRY",
                        value / exchange_housing$usdtry[1],
                        value / exchange_housing$kfe_nominal[1])
  ) |>
  ggplot(aes(period, value_norm, colour = series, group = series)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  scale_colour_manual(values = c(pal$secondary, pal$accent), name = NULL) +
  scale_y_continuous(labels = label_number(suffix = "×")) +
  labs(
    title    = "USD/TRY vs. KFE — both normalised to Q1 2018 = 1×",
    subtitle = "Currency depreciation consistently leads or accompanies KFE spikes",
    x = NULL, y = "Index (Q1 2018 = 1×)",
    caption  = "Source: TCMB EVDS (USD/TRY); TCMB KFE."
  ) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 7))
```

The two series track closely — exchange rate acceleration consistently **leads or
coincides with KFE acceleration**. The 2021 Q4 TRY collapse (from ~8.5 to 13+ USD/TRY)
directly preceded the 2022 nominal KFE explosion of ~183% YoY.

### 4.2 YoY change correlation

```{r fx-correlation, fig.height=4}
exchange_housing |>
  filter(!is.na(kfe_yoy_pct)) |>
  mutate(
    usdtry_yoy = (usdtry / lag(usdtry, 4) - 1) * 100
  ) |>
  filter(!is.na(usdtry_yoy)) |>
  ggplot(aes(usdtry_yoy, kfe_yoy_pct)) +
  geom_point(aes(colour = factor(year)), size = 3) +
  geom_smooth(method = "lm", se = TRUE,
              colour = pal$accent, fill = pal$accent, alpha = 0.1,
              linewidth = 0.8) +
  scale_colour_viridis_d(name = "Year", option = "D") +
  scale_x_continuous(labels = label_percent(scale = 1)) +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  labs(
    title    = "USD/TRY annual change vs. KFE annual change — quarterly data 2019–2024",
    subtitle = "Strong positive correlation: currency depreciation feeds into housing prices",
    x = "USD/TRY YoY change (%)", y = "KFE nominal YoY change (%)",
    caption  = "Source: TCMB EVDS; TCMB KFE. Each point = one quarter."
  )
```

### 4.3 Real vs. nominal KFE growth — separating inflation premium

```{r real-premium, fig.height=4}
exchange_housing |>
  filter(!is.na(kfe_yoy_pct)) |>
  select(period, year, `Nominal KFE YoY` = kfe_yoy_pct,
         `CPI YoY` = cpi_yoy_pct, `Real KFE YoY` = kfe_real_yoy_pct) |>
  pivot_longer(-c(period, year), names_to = "series", values_to = "pct") |>
  ggplot(aes(period, pct, colour = series, group = series)) +
  geom_line(linewidth = 0.9) +
  geom_hline(yintercept = 0, linetype = "dashed",
             colour = pal$muted, linewidth = 0.4) +
  scale_colour_manual(
    values = c(pal$muted, pal$accent, pal$secondary),
    name   = NULL
  ) +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  labs(
    title    = "Nominal KFE, CPI, and real KFE change — quarterly, 2019–2024",
    subtitle = "Real KFE has been negative in most quarters since Q3 2022",
    x = NULL, y = "YoY change (%)",
    caption  = "Source: TCMB KFE; TURKSTAT CPI."
  ) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 7))
```

---

## 5. Supply-Side: Sales and Building Permits

### 5.1 Housing sales — total and by type

```{r sales-trend, fig.height=4.5}
sales_permits |>
  select(year, `Total sales` = sales_total,
         `First-hand` = sales_first_hand,
         `Second-hand` = sales_second_hand) |>
  filter(year >= 2015) |>
  pivot_longer(-year, names_to = "type", values_to = "count") |>
  filter(!is.na(count)) |>
  ggplot(aes(year, count, colour = type)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_colour_manual(
    values = c(pal$accent, pal$secondary, pal$warn),
    name   = NULL
  ) +
  scale_x_continuous(breaks = 2015:2024) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "Housing sales by type, 2015–2024",
    subtitle = "Total sales peaked in 2020 (credit stimulus) and have since declined",
    x = NULL, y = "Number of units sold",
    caption  = "Source: TÜİK İnşaat ve Konut İstatistikleri."
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Sales peaked in **2020** at 1.5 million units, boosted by the government's subsidised
mortgage campaign (June 2020). Since then, rising prices and tightening credit have
driven sales down to **1.07 million** in 2024 — the lowest since 2010.

### 5.2 Building permits — the supply pipeline

```{r permits-trend, fig.height=4}
sales_permits |>
  select(year, `Building permits (units)` = permits_residential_units,
         `Total sales` = sales_total) |>
  pivot_longer(-year, names_to = "series", values_to = "value") |>
  ggplot(aes(year, value, colour = series)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.2) +
  scale_colour_manual(values = c(pal$accent, pal$secondary), name = NULL) +
  scale_x_continuous(breaks = 2010:2024) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "Building permits vs. housing sales, 2010–2024",
    subtitle = "Supply pipeline (permits) halved from 2015 peak — structural undersupply ahead",
    x = NULL, y = "Number of units",
    caption  = "Source: TÜİK İnşaat ve Konut İstatistikleri."
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Building permits — the leading indicator of future supply — fell from **1.22 million
units** in 2015 to **549,000** in 2024, a **55% decline**. Given the 2–4 year lag
between permit issuance and delivery, this signals a structural supply shortfall through
at least 2026–2027, which will continue to support nominal prices despite real declines.

### 5.3 Interactive full data table

```{r sales-dt}
sales_permits |>
  mutate(
    sales_total       = comma(sales_total),
    permits_residential_units = comma(permits_residential_units),
    second_hand_share = paste0(round(second_hand_share, 1), "%"),
    mortgage_share    = ifelse(is.na(mortgage_share), "—",
                               paste0(round(mortgage_share, 1), "%"))
  ) |>
  select(year, sales_total, permits_residential_units,
         second_hand_share, mortgage_share, sales_foreign) |>
  datatable(rownames = FALSE, filter = "top",
            options  = list(pageLength = 15, dom = "ftp"),
            caption  = "Annual housing market supply-demand indicators, 2010–2024")
```

---

## Conclusions

Four structural findings emerge from this analysis:

1. **The Turkish housing boom is largely nominal, not real.** KFE rose ~48× nominally
   but only ~18% in real terms since 2010, and that real gain has partially reversed
   since 2022. Buyers in 2021–2023 effectively paid an inflation premium.

2. **Affordability has collapsed to crisis levels.** The 3× worsening in the
   months-of-minimum-wage metric is structural, not cyclical — it reflects a decade
   of supply under-investment coinciding with income growth that could not keep pace
   with price inflation or currency depreciation.

3. **Exchange rate risk is the dominant driver.** The pass-through from USD/TRY
   depreciation to nominal housing prices is rapid and strong — Turkish housing has
   effectively acted as a partial inflation hedge in lira terms, drawing in domestic
   savings as a store of value.

4. **The supply pipeline threatens prolonged undersupply.** The 55% fall in building
   permits since 2015, combined with post-earthquake relocation demand, means structural
   supply constraints will persist even if nominal price growth slows further. True
   affordability improvement requires both price cooling and a supply-side policy response.
