---
title: "Analysis"
subtitle: "A decade of road traffic risk in Türkiye, 2015–2024"
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(scales)
library(knitr)
library(tidyr)
library(forcats)
library(stringr)
library(DT)

load("data/traffic_accidents_tr.RData")

pal <- list(
  primary   = "#1a1814",
  secondary = "#2c3e50",
  accent    = "#c0392b",
  good      = "#27ae60",
  warn      = "#e67e22",
  muted     = "#95a5a6"
)

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title       = element_text(face = "bold", size = 13, hjust = 0),
      plot.subtitle    = element_text(colour = "#6b6560", size = 10.5, hjust = 0),
      plot.caption     = element_text(colour = "#95a5a6", size = 8, hjust = 0),
      plot.background  = element_rect(fill = "#f8f5f0", colour = NA),
      panel.background = element_rect(fill = "#f8f5f0", colour = NA),
      panel.grid.major = element_line(colour = "#e0dbd3"),
      panel.grid.minor = element_blank(),
      legend.position  = "top",
      legend.text      = element_text(size = 9),
      axis.text        = element_text(colour = "#6b6560", size = 9),
      axis.title       = element_text(colour = "#1a1814", size = 10)
    )
)
```

::: {.callout-note icon="false"}
### Key Takeaways

- **Volume up, per-vehicle risk down.** Total accidents rose 46% (2015→2024) while
  deaths per 100,000 vehicles fell from 37.7 to 20.3.
- **2020 broke the trend.** Pandemic restrictions cut deaths to 4,866 (decade low);
  by 2023 both accidents and deaths overshot pre-pandemic levels.
- **Risk is concentrated.** Driver fault: ~88–90% of causes every year. Top single
  error: speed-related (40.8% of driver faults). Motorcycles: 31.5% of involvements,
  40.0% of driver deaths.
- **Urban ≠ most dangerous.** 86% of fatal/injury accidents are in cities but only
  56% of deaths — rural fatality risk per crash is ~4× higher.
- **Türkiye is a European outlier.** ~74 deaths per million population vs. EU-27
  average of ~47.
:::

---

## 1. Trend Analysis, 2015–2024

### 1.1 Headline indicators

```{r trend-facet, fig.height=5.5}
annual |>
  select(year, total_accidents, fatal_injury_accidents, deaths_total, injured) |>
  pivot_longer(-year, names_to = "metric", values_to = "value") |>
  mutate(metric = recode(metric,
    total_accidents        = "Total accidents",
    fatal_injury_accidents = "Fatal/injury accidents",
    deaths_total           = "Deaths",
    injured                = "Injured"
  )) |>
  ggplot(aes(year, value)) +
  geom_line(colour = pal$primary, linewidth = 0.9) +
  geom_point(colour = pal$accent, size = 2.2) +
  facet_wrap(~ metric, scales = "free_y", ncol = 2) +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(breaks = 2015:2024) +
  labs(
    title    = "Headline trends, 2015–2024",
    subtitle = "Each panel has its own y-axis — shape reveals the story, not absolute values",
    x = NULL, y = NULL,
    caption  = "Source: TÜİK Road Traffic Accident Statistics, 2015–2024."
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

All four indicators fell sharply in 2020 then rebounded. The key asymmetry:
**deaths grew more slowly than accidents**, reflecting safer outcomes per crash
through better vehicle safety, faster emergency response, and road improvements.

### 1.2 Normalised fatality rates

```{r rates-plot, fig.height=4}
rates |>
  select(year,
         `Per 100k population` = deaths_per_100k_population,
         `Per 100k vehicles`   = deaths_per_100k_vehicles) |>
  pivot_longer(-year, names_to = "metric", values_to = "value") |>
  ggplot(aes(year, value, colour = metric)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5) +
  scale_colour_manual(values = c(pal$accent, pal$secondary), name = NULL) +
  scale_x_continuous(breaks = 2015:2024) +
  labs(
    title    = "Fatality rates normalised for exposure, 2015–2024",
    subtitle = "Both series trend downward — Türkiye is getting safer per vehicle driven",
    x = NULL, y = "Deaths",
    caption  = "Source: TÜİK. 2024 vehicle-km not yet published."
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Per-vehicle fatalities fell **46%** (37.7 → 20.3). Per-capita fell **23%** (9.6 → 7.4).
The larger per-vehicle improvement reflects that the fleet grew faster than the
population, so each car is becoming safer faster than overall mortality risk is falling.

### 1.3 Fleet growth vs. deaths — decoupling

```{r decoupling, fig.height=4.2}
ggplot(annual, aes(registered_vehicles / 1e6, deaths_total)) +
  geom_path(colour = pal$muted, linewidth = 0.8,
            arrow = arrow(length = unit(0.2, "cm"), type = "closed")) +
  geom_point(aes(colour = year == 2020), size = 3) +
  geom_text(aes(label = year), vjust = -1.1, size = 3, colour = pal$primary) +
  scale_colour_manual(
    values = c(`FALSE` = pal$secondary, `TRUE` = pal$accent),
    guide  = "none"
  ) +
  scale_x_continuous(labels = label_number(suffix = "M")) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "More vehicles, fewer deaths — a decade of decoupling",
    subtitle = "Arrow traces 2015 → 2024; red point = 2020 pandemic outlier",
    x = "Registered vehicles", y = "Deaths",
    caption  = "Source: TÜİK."
  )
```

The path arcs **down and to the right** — the classic signature of a system gaining
safety per unit of exposure. Every year (except the post-2020 rebound) sits lower
on the y-axis than the same fleet-size point would have in an earlier year.

---

## 2. Causal Factors

### 2.1 Driver vs. other causes over time

```{r fault-time, fig.height=4}
fault_pct |>
  pivot_longer(-year, names_to = "factor", values_to = "share") |>
  mutate(factor = recode(factor,
    driver_pct     = "Driver",
    pedestrian_pct = "Pedestrian",
    passenger_pct  = "Passenger",
    vehicle_pct    = "Vehicle defect",
    road_pct       = "Road defect"
  )) |>
  filter(factor != "Driver") |>
  ggplot(aes(year, share, colour = factor)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_colour_manual(
    values = c(pal$accent, pal$good, pal$warn, pal$secondary),
    name   = NULL
  ) +
  scale_x_continuous(breaks = 2015:2024) +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  labs(
    title    = "Non-driver causes, 2015–2024",
    subtitle = "Driver fault (86–90% each year) is excluded so smaller categories are readable",
    x = NULL, y = "Share of attributable causes",
    caption  = "Source: KGM 2024, Tablo 1.3."
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Driver error is dominant and remarkably stable at 86–90% every year. Road defects
are minor and falling, consistent with continued infrastructure investment.

### 2.2 Top driver errors in 2024

```{r driver-errors, fig.height=5}
driver_faults |>
  arrange(desc(fault_count)) |>
  slice_head(n = 10) |>
  mutate(
    label = str_wrap(fault_type, 48),
    label = fct_reorder(label, fault_count)
  ) |>
  ggplot(aes(fault_count, label)) +
  geom_col(fill = pal$secondary) +
  geom_text(
    aes(label = paste0(comma(fault_count), "  (", fault_pct, "%)")),
    hjust = -0.05, size = 2.9
  ) +
  scale_x_continuous(labels = comma, expand = expansion(mult = c(0, 0.3))) +
  labs(
    title    = "Top 10 driver-error categories, 2024",
    subtitle = "KGM road network — 61,459 attributable driver faults total",
    x = "Fault count", y = NULL,
    caption  = "Source: KGM 2024, Tablo 2.12."
  )
```

The top three errors — speeding (40.8%), lane violations (17.9%), rear-end (15.5%) —
together explain **74%** of attributable driver faults. Enforcement concentrated on
these three behaviours could address a large share of the burden efficiently.

---

## 3. Vehicle Type Risk

### 3.1 Involvement vs. deaths

```{r vehicle-disproportion, fig.height=5}
vehicles |>
  filter(vehicles_involved >= 1000) |>
  mutate(vehicle_type = fct_reorder(vehicle_type, vehicles_involved)) |>
  select(vehicle_type,
         `Vehicles involved %` = vehicles_pct,
         `Driver deaths %`     = deaths_pct,
         `Injured drivers %`   = injured_pct) |>
  pivot_longer(-vehicle_type) |>
  ggplot(aes(value, vehicle_type, fill = name)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.75) +
  scale_fill_manual(
    values = c(pal$secondary, pal$accent, pal$warn),
    name   = NULL
  ) +
  scale_x_continuous(labels = label_percent(scale = 1)) +
  labs(
    title    = "Disproportionate fatality risk by vehicle type, 2024",
    subtitle = "Red bar > blue bar = over-represented in deaths relative to crash share",
    x = "Share of national total", y = NULL,
    caption  = "Source: TÜİK 2024, Tablo 1.4. Types with < 1,000 involvements omitted."
  )
```

Motorcycles are 31.5% of involvements yet **40.0%** of driver deaths and **54.1%**
of driver injuries. Cars are the reverse — 41.8% of involvements, 35.2% of deaths —
thanks to structural passive safety.

### 3.2 Deaths per 1,000 vehicles involved

```{r vehicle-rate, fig.height=4.5}
vehicles |>
  filter(vehicles_involved >= 500) |>
  mutate(vehicle_type = fct_reorder(vehicle_type, deaths_per_1000_involved)) |>
  ggplot(aes(deaths_per_1000_involved, vehicle_type)) +
  geom_col(fill = pal$accent) +
  geom_text(
    aes(label = round(deaths_per_1000_involved, 1)),
    hjust = -0.15, size = 3
  ) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(
    title    = "Driver deaths per 1,000 vehicles involved, 2024",
    subtitle = "Controls for crash frequency — pure risk-per-incident view",
    x = "Deaths per 1,000 involved vehicles", y = NULL,
    caption  = "Source: TÜİK 2024, Tablo 1.4."
  )
```

Tractors: ~46 driver deaths per 1,000 involvements. Trucks: ~14. Motorcycles: ~10.
Cars: ~6.6. Structural protection — not just speed — determines lethality.

---

## 4. When and Where

### 4.1 Monthly pattern

```{r monthly, fig.height=4}
monthly |>
  mutate(month = factor(month, levels = monthly$month)) |>
  ggplot(aes(month, accidents, fill = severity_index)) +
  geom_col() +
  geom_text(aes(label = comma(accidents)), vjust = -0.4, size = 2.7) +
  scale_fill_gradient(
    low  = pal$secondary,
    high = pal$accent,
    name = "Deaths / 100 accidents"
  ) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.1))) +
  labs(
    title    = "Monthly accident counts, KGM network, 2024",
    subtitle = "Bars coloured by severity — August peaks in both volume and lethality",
    x = NULL, y = NULL,
    caption  = "Source: KGM 2024, Tablo 2.6."
  ) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

August accounts for **11.3%** of annual accidents (2.1× February's 5.4%). Holiday
traffic density and higher intercity speeds drive both volume and severity peaks.

### 4.2 Day of week

```{r weekday, fig.height=3.5}
weekday |>
  mutate(day = factor(day, levels = weekday$day)) |>
  ggplot(aes(day, accidents)) +
  geom_col(fill = pal$secondary) +
  geom_text(aes(label = comma(accidents)), vjust = -0.4, size = 3) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.1))) +
  labs(
    title    = "Accidents by day of week, 2024",
    subtitle = "Saturday peaks (+12% vs Monday), consistent with weekend travel",
    x = NULL, y = NULL,
    caption  = "Source: KGM 2024, Tablo 2.7."
  )
```

### 4.3 Daylight severity

```{r daylight, fig.height=3}
daylight |>
  mutate(light_condition = fct_reorder(light_condition, severity_index)) |>
  ggplot(aes(severity_index, light_condition)) +
  geom_col(fill = pal$accent) +
  geom_text(
    aes(label = paste0(round(severity_index, 2), " deaths / 100 acc.")),
    hjust = -0.05, size = 3.2
  ) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.35))) +
  labs(
    title    = "Fatality severity by daylight condition, 2024",
    subtitle = "Night-time accidents are ~38% more lethal than daytime",
    x = NULL, y = NULL,
    caption  = "Source: KGM 2024, Tablo 2.8."
  )
```

Night accidents kill at **7.6 per 100 accidents** vs 5.5 in daylight — a 38%
relative increase. Reduced visibility, driver fatigue, and higher truck share at
night all contribute.

### 4.4 Urban vs. rural

```{r settlement, fig.height=3}
settlement |>
  mutate(
    share_acc    = fatal_injury_accidents / sum(fatal_injury_accidents),
    share_deaths = deaths_total           / sum(deaths_total)
  ) |>
  select(settlement,
         `Fatal/injury accidents` = share_acc,
         `Deaths`                 = share_deaths) |>
  pivot_longer(-settlement) |>
  ggplot(aes(name, value, fill = settlement)) +
  geom_col(width = 0.5) +
  geom_text(
    aes(label = percent(value, 0.1)),
    position  = position_stack(vjust = 0.5),
    colour    = "white", size = 4, fontface = "bold"
  ) +
  scale_y_continuous(labels = label_percent()) +
  scale_fill_manual(values = c(pal$secondary, pal$accent), name = NULL) +
  coord_flip() +
  labs(
    title    = "Urban vs. rural, 2024",
    subtitle = "Cities: 86% of crashes, 56% of deaths — rural fatality risk is ~4× higher per crash",
    x = NULL, y = NULL,
    caption  = "Source: TÜİK 2024, Tablo 1.2."
  )
```

### 4.5 Crash type severity

```{r crash-type, fig.height=5.5}
accident_types |>
  filter(accidents >= 200) |>
  mutate(
    atype = str_wrap(accident_type, 30),
    atype = fct_reorder(atype, fatality_rate_pct)
  ) |>
  ggplot(aes(fatality_rate_pct, atype)) +
  geom_col(aes(fill = accidents)) +
  geom_text(
    aes(label = paste0(round(fatality_rate_pct, 1), "% (n=", comma(accidents), ")")),
    hjust = -0.05, size = 2.7
  ) +
  scale_fill_gradient(
    low  = pal$muted,
    high = pal$primary,
    name = "Accident count", labels = comma
  ) +
  scale_x_continuous(
    labels = label_percent(scale = 1),
    expand = expansion(mult = c(0, 0.4))
  ) +
  labs(
    title    = "Fatality rate by crash type, 2024",
    subtitle = "Pedestrian strikes: ~20 deaths / 100 crashes — 4× the network average",
    x = "Deaths per 100 accidents", y = NULL,
    caption  = "Source: KGM 2024, Tablo 2.4."
  )
```

**Pedestrian strikes** dominate severity despite low volume. Volume-leaders
(lane-departure, side-impact, rear-end) account for most *total* deaths through
sheer frequency — both severity and volume dimensions matter for policy.

---

## 5. European Comparison

### 5.1 Deaths per million population

```{r eu-bar, fig.height=6.5}
eu_comparison |>
  mutate(
    country = fct_reorder(country, deaths_per_million_population),
    is_tr   = country == "Turkiye"
  ) |>
  ggplot(aes(deaths_per_million_population, country, fill = is_tr)) +
  geom_col() +
  geom_text(
    aes(label = deaths_per_million_population),
    hjust = -0.2, size = 2.7
  ) +
  scale_fill_manual(
    values = c(`FALSE` = pal$secondary, `TRUE` = pal$accent),
    guide  = "none"
  ) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title    = "Road deaths per million population, 2024",
    subtitle = "Türkiye (red) vs. EU-27. EU figures reflect 2022 data.",
    x = NULL, y = NULL,
    caption  = "Source: TÜİK 2024, Tablo 1.6 (EU via Transport Statistical Pocketbook 2024)."
  ) +
  theme(axis.text.y = element_text(size = 9))
```

Türkiye's **74 deaths per million** places it in the upper quarter of the European
distribution — comparable to Croatia (71), well above Germany (33) and France (48).

### 5.2 Motorisation vs. fatality burden

```{r eu-scatter, fig.height=5}
eu_comparison |>
  mutate(is_tr = country == "Turkiye") |>
  ggplot(aes(cars_per_1000, deaths_per_million_population)) +
  geom_smooth(method = "lm", se = FALSE,
              colour = pal$muted, linewidth = 0.7) +
  geom_point(aes(colour = is_tr), size = 3) +
  geom_text(
    aes(label = country, colour = is_tr),
    vjust = -1, size = 2.5, show.legend = FALSE
  ) +
  scale_colour_manual(
    values = c(`FALSE` = pal$secondary, `TRUE` = pal$accent),
    guide  = "none"
  ) +
  labs(
    title    = "Higher motorisation → lower per-capita road deaths (Europe, 2024)",
    subtitle = "Türkiye (red): low motorisation AND high fatalities — a structural outlier",
    x = "Cars per 1,000 population", y = "Deaths per million population",
    caption  = "Source: TÜİK 2024, Tablo 1.6."
  )
```

Across Europe there is a clear **negative association** between motorisation and
per-capita fatalities. Türkiye sits bottom-left — 189 cars/1,000 vs. EU average
540+ — yet with high fatalities, indicating the gap is structural, not mechanical.

---

## 6. Full Data Summary

```{r full-summary-dt}
annual |>
  mutate(
    fatal_injury_share  = percent(fatal_injury_share, 0.1),
    deaths_per_accident = round(deaths_per_accident, 4),
    registered_vehicles = comma(registered_vehicles),
    total_accidents     = comma(total_accidents),
    deaths_total        = comma(deaths_total),
    injured             = comma(injured)
  ) |>
  datatable(
    rownames = FALSE,
    filter   = "top",
    options  = list(pageLength = 10, dom = "ftp"),
    caption  = "Annual headline series, 2015–2024 (interactive)"
  )
```

---

## Conclusions

Three findings carry direct policy weight:

1. **The normalised trend improved over the decade but reversed after 2021.**
   Fleet growth has recently outpaced safety improvement; the per-capita gap
   to Europe is closing only slowly.

2. **A small set of behaviours explains a large share of harm.** Speed-related
   errors account for >40% of driver faults; pedestrian strikes are ~4% of crashes
   but a disproportionate share of fatalities. Targeted enforcement and pedestrian
   infrastructure are high-leverage interventions.

3. **Vulnerability is type-bound.** Motorcyclists and tractor operators carry
   4–7× the per-crash fatality risk of car drivers; night-time and rural accidents
   are markedly deadlier per event. Protective-gear enforcement, rural road upgrades,
   and visibility improvements each address structurally distinct slices of the problem.
