---
title: "Data"
subtitle: "Sources, preprocessing pipeline, and the unified dataset"
---

```{r setup, include=FALSE}
library(dplyr)
library(knitr)
library(scales)
library(DT)
load("data/housing_market_tr.RData")
```

## Data Sources

We integrate four official Turkish data sources:

| # | Source | Series | Coverage |
|---|---|---|---|
| 1 | TCMB Konut Fiyat Endeksi | Nominal and real KFE, NUTS-2 regional KFE | Monthly, 2010-2024 |
| 2 | TCMB EVDS -- Consumer Prices | CPI index for deflation | Monthly, 2010-2024 |
| 3 | TCMB EVDS -- Exchange Rates | USD/TRY monthly average | Monthly, 2010-2024 |
| 4 | TUIK Insaat ve Konut Istatistikleri | Housing sales, building permits | Annual, 2010-2024 |

**Direct links:**

- TCMB KFE: <https://www.tcmb.gov.tr/wps/wcm/connect/tr/tcmb+tr/main+menu/istatistikler/reel+sektor+istatistikleri/konut+fiyat+endeksi/>
- TCMB EVDS (CPI, FX): <https://evds2.tcmb.gov.tr>
- TUIK Housing Statistics: <https://data.tuik.gov.tr/Kategori/GetKategori?p=insaat-ve-konut-116>

::: {.callout-note}
**Why this dataset?** Turkey's housing market has undergone one of the world's most
extreme price cycles in 2020-2024, driven by a unique combination of currency collapse,
triple-digit inflation, post-earthquake displacement, and supply-side contraction.
This makes it an exceptionally rich case for understanding how macroeconomic shocks
transmit into asset markets.
:::

---

## Preprocessing Pipeline

Raw TCMB and TUIK data require aggregation and cleaning. The full pipeline is in
`R/01_preprocess.R`. Key steps:

**Step 1 -- Manual extraction.**
Series were extracted from TCMB publications and TUIK statistical tables into
`data/raw_*.csv` files, preserving original source values.

**Step 2 -- Tidying.**
All CSV files are read with UTF-8 encoding and reshaped to one-row-per-observation
with consistent snake_case column names.

**Step 3 -- Derived variables.**

| Variable | Object | Formula |
|---|---|---|
| kfe_nominal_yoy | annual_macro | (kfe_nominal / lag - 1) x 100 |
| kfe_real_yoy | annual_macro | (kfe_real / lag - 1) x 100 |
| cum_2020_2024_pct | regional_nuts2 | Compounded 5-year nominal change |
| months_100m2 | affordability | already in source data |

```{r preprocess-code, eval=FALSE}
# R/01_preprocess.R -- run once before rendering
library(readr); library(dplyr)

rd <- function(f) read_csv(file.path("data", f),
                            show_col_types = FALSE,
                            locale = locale(encoding = "UTF-8"))

monthly_data <- rd("raw_monthly_kfe_cpi_usdtry.csv")

annual_macro <- rd("raw_annual_macro.csv") |>
  mutate(
    kfe_nominal_yoy = (kfe_nominal_2010base / lag(kfe_nominal_2010base) - 1) * 100,
    kfe_real_yoy    = (kfe_real_2010base    / lag(kfe_real_2010base)    - 1) * 100
  )

regional_nuts2 <- rd("raw_regional_nuts2.csv") |>
  mutate(
    cum_2020_2024_pct =
      ((1 + kfe_2020_yoy_pct/100) * (1 + kfe_2021_yoy_pct/100) *
       (1 + kfe_2022_yoy_pct/100) * (1 + kfe_2023_yoy_pct/100) *
       (1 + kfe_2024_yoy_pct/100) - 1) * 100
  )

city_monthly   <- rd("raw_city_kfe_monthly.csv")
affordability  <- rd("raw_affordability_annual.csv")
quarterly_fx   <- rd("raw_quarterly_kfe_fx.csv")
housing_sales  <- rd("raw_housing_sales_annual.csv")
permits        <- rd("raw_building_permits_annual.csv")
minwage        <- rd("raw_minimum_wage_history.csv")
kfe_new_exist  <- rd("raw_kfe_new_vs_existing.csv")

save(monthly_data, annual_macro, regional_nuts2, city_monthly,
     affordability, quarterly_fx, housing_sales, permits,
     minwage, kfe_new_exist,
     file = "data/housing_market_tr.RData")
```

---

## The Final Dataset

**[Download housing_market_tr.RData](data/housing_market_tr.RData)**

```r
load("data/housing_market_tr.RData")  # loads 10 objects
```

### Contents

```{r contents-table}
tibble::tibble(
  Object = c("annual_macro", "monthly_data", "regional_nuts2", "city_monthly",
             "affordability", "quarterly_fx", "housing_sales", "permits",
             "minwage", "kfe_new_exist"),
  Description = c(
    "Annual KFE (nominal+real), CPI, USD/TRY, sales, permits 2010-2024 (17 cols)",
    "Monthly KFE, CPI, USD/TRY time series 2010-2024 (180 rows)",
    "26 NUTS-2 regions: annual KFE change % and 2024 median price",
    "Istanbul, Ankara, Izmir monthly KFE + m2 price comparison",
    "Annual affordability: m2 price, min. wage, months-to-buy 2010-2024",
    "Quarterly USD/TRY vs KFE for correlation analysis 2015-2024",
    "Annual housing sales by type (total, first-hand, mortgage, foreign)",
    "Annual building permits and occupancy certificates 2010-2024",
    "Minimum wage gross/net history with mid-year adjustments 2010-2024",
    "New (YKFE) vs existing (YOKFE) housing price index 2010-2024"
  ),
  Rows = c(
    nrow(annual_macro), nrow(monthly_data), nrow(regional_nuts2),
    nrow(city_monthly), nrow(affordability), nrow(quarterly_fx),
    nrow(housing_sales), nrow(permits), nrow(minwage), nrow(kfe_new_exist)
  )
) |> kable()
```

### Preview -- annual_macro

```{r dt-annual}
annual_macro |>
  select(year, kfe_nominal_2010base, kfe_real_2010base, cpi_yoy_pct,
         usdtry_annual_avg, kfe_nominal_yoy, kfe_real_yoy) |>
  mutate(
    kfe_nominal_2010base = round(kfe_nominal_2010base, 1),
    kfe_real_2010base    = round(kfe_real_2010base, 1),
    cpi_yoy_pct          = round(cpi_yoy_pct, 2),
    kfe_nominal_yoy      = round(kfe_nominal_yoy, 1),
    kfe_real_yoy         = round(kfe_real_yoy, 1)
  ) |>
  datatable(
    rownames = FALSE,
    options  = list(pageLength = 15, dom = "t"),
    caption  = "Annual KFE + macro series, 2010-2024"
  )
```

### Preview -- regional_nuts2 (26 NUTS-2 regions)

```{r dt-regional}
regional_nuts2 |>
  select(region_code, region_name, kfe_2022_yoy_pct, kfe_2023_yoy_pct,
         kfe_2024_yoy_pct, cum_2020_2024_pct, median_price_try_m2_2024) |>
  mutate(
    cum_2020_2024_pct        = round(cum_2020_2024_pct, 1),
    median_price_try_m2_2024 = comma(median_price_try_m2_2024)
  ) |>
  datatable(
    rownames = FALSE,
    filter   = "top",
    options  = list(pageLength = 13, dom = "ftp"),
    caption  = "Regional KFE change % and 2024 median unit price (26 NUTS-2 regions)"
  )
```

### Preview -- monthly_data (180 months)

```{r dt-monthly}
monthly_data |>
  filter(year >= 2020) |>
  mutate(
    kfe_nominal_2010base = round(kfe_nominal_2010base, 1),
    cpi_yoy_pct          = round(cpi_yoy_pct, 2)
  ) |>
  datatable(
    rownames = FALSE,
    filter   = "top",
    options  = list(pageLength = 10, dom = "ftp"),
    caption  = "Monthly KFE + CPI + USD/TRY (showing 2020-2024; full dataset has 180 rows)"
  )
```

---

## Data Quality Notes

- **KFE methodology revision (July 2024).** TCMB revised the KFE from a quarterly
  to a monthly hedonic regression model, rebasing to 2023=100. The monthly series
  here preserves both bases (2023=100 and 2010=100) for comparability.

- **Regional data.** NUTS-2 regional breakdowns (26 regions) are published as annual
  % change figures. Median unit prices are TCMB quarterly figures averaged to annual.

- **Sales data.** First-hand / second-hand split is available from 2015. 2024 total
  reflects TUIK revised figure of 1,319,539.

- **CPI deflation.** Real KFE uses TURKSTAT official CPI (2003=100 base, consistent
  series converted to 2010=100 for this analysis).

- **Missing values.** Some regional 2024 figures are NA where TCMB had not published
  quarterly data by the time of data collection. These are coded as NA, not imputed.
