---
title: "Data"
subtitle: "Sources, preprocessing pipeline, and the unified dataset"
---

```{r setup, include=FALSE}
library(dplyr)
library(knitr)
library(scales)
library(DT)

load("data/traffic_accidents_tr.RData")
```

## Data Sources

We combine three official Turkish open-data sources into a single tidy collection:

| Source | Coverage | Used for |
|---|---|---|
| **TÜİK** — *Karayolu Trafik Kaza İstatistikleri 2024* | National totals, vehicle types, EU comparison, urban/rural | Annual time series 2015–2024 |
| **KGM** — *2024 Trafik Kazalarına Ait Özet Bilgiler* | KGM road network (highways, state, provincial roads) | Monthly, weekday, daylight, fault factors |
| **EGM / JGK** | Underlying records aggregated into KGM/TÜİK tables | Via upstream tables |

**Direct links:**

- TÜİK 2024 bulletin: <https://data.tuik.gov.tr/Bulten/Index?p=Road-Traffic-Accident-Statistics-2024-54056>
- KGM 2024 summary PDF: <https://www.kgm.gov.tr/SiteCollectionDocuments/KGMdocuments/Trafik/Trafik-kaza-ozetbilgi.pdf>
- EGM Trafik Başkanlığı: <https://www.trafik.gov.tr/istatistikler37>

::: {.callout-note}
**Why this dataset?** Traffic safety in Türkiye is a high-stakes public-interest
topic with rich official data, decade-long temporal coverage, vehicle-type and
causal-factor granularity, and a ready European benchmark — enabling descriptive,
comparative, and causal analyses in a single coherent narrative.
:::

---

## Preprocessing Pipeline

Raw TÜİK/KGM publications are not available as analysis-ready files. Our three-step pipeline is fully reproducible from `R/01_preprocess.R`:

**Step 1 — Manual extraction.**  
Each table was extracted from official PDFs and saved as `data/raw_*.csv`. Separating
transcription from analysis makes corrections trivial.

**Step 2 — Tidying in R.**  
CSV files are read with UTF-8 encoding, percentage columns converted to numeric, and
tables reshaped to one-row-per-observation with `snake_case` column names.

**Step 3 — Derived variables.**

| Variable | Object | Formula |
|---|---|---|
| `fatal_injury_share` | `annual` | fatal_injury_accidents / total_accidents |
| `deaths_per_accident` | `annual` | deaths_total / fatal_injury_accidents |
| `severity_index` | `monthly`, `weekday`, `daylight` | deaths / accidents × 100 |
| `deaths_per_1000_involved` | `vehicles` | driver_deaths_total / vehicles_involved × 1000 |
| `fatality_rate_pct` | `accident_types` | deaths / accidents × 100 |

```{r preprocess-show, eval=FALSE}
# R/01_preprocess.R  —  run once before rendering
# setwd("sansasyonel/")  then  source("R/01_preprocess.R")

library(readr); library(dplyr)

read_raw <- function(f) read_csv(file.path("data", f),
                                  show_col_types = FALSE,
                                  locale = locale(encoding = "UTF-8"))

annual <- read_raw("raw_annual_2015_2024.csv") |>
  mutate(fatal_injury_share  = fatal_injury_accidents / total_accidents,
         deaths_per_accident = deaths_total / fatal_injury_accidents)

fault_pct <- read_raw("raw_fault_pct_2015_2024.csv")

vehicles <- read_raw("raw_vehicle_type_2024.csv") |>
  mutate(deaths_per_1000_involved = 1000 * driver_deaths_total / vehicles_involved)

rates    <- read_raw("raw_rates_2015_2024.csv")
monthly  <- read_raw("raw_monthly_2024.csv")   |> mutate(severity_index = deaths / accidents * 100)
weekday  <- read_raw("raw_dayofweek_2024.csv") |> mutate(severity_index = deaths / accidents * 100)
daylight <- read_raw("raw_daylight_2024.csv")  |> mutate(severity_index = deaths / accidents * 100)

driver_faults  <- read_raw("raw_driver_faults_2024.csv")
accident_types <- read_raw("raw_accident_type_2024.csv") |>
  mutate(fatality_rate_pct = deaths / accidents * 100)
eu_comparison  <- read_raw("raw_eu_comparison_2024.csv")
settlement     <- read_raw("raw_settlement_2024.csv")

save(annual, fault_pct, vehicles, rates, monthly, weekday,
     daylight, driver_faults, accident_types, eu_comparison, settlement,
     file = "data/traffic_accidents_tr.RData")
```

---

## The Final Dataset

All analysis pages load one file:

**[⬇ Download traffic_accidents_tr.RData](data/traffic_accidents_tr.RData)**

```r
load("data/traffic_accidents_tr.RData")  # 11 data frame available
```

### Contents

```{r contents-table}
tibble::tibble(
  Object = c("annual","fault_pct","vehicles","rates",
             "monthly","weekday","daylight","driver_faults",
             "accident_types","eu_comparison","settlement"),
  Description = c(
    "National annual totals 2015–2024",
    "Yearly share of accident causes by category",
    "Vehicle-type involvement and driver outcomes, 2024",
    "Per-100k-population / per-100k-vehicle rates, 2015–2024",
    "Monthly accident/death/injured counts, KGM network, 2024",
    "Day-of-week distribution, KGM network, 2024",
    "Daylight-condition distribution, KGM network, 2024",
    "Top driver fault categories, KGM network, 2024",
    "Accident occurrence types, KGM network, 2024",
    "Cross-country comparison with EU-27, 2024",
    "Urban vs. rural breakdown, 2024"
  ),
  Rows = c(nrow(annual), nrow(fault_pct), nrow(vehicles), nrow(rates),
           nrow(monthly), nrow(weekday), nrow(daylight), nrow(driver_faults),
           nrow(accident_types), nrow(eu_comparison), nrow(settlement))
) |> kable()
```

### Interactive preview — `annual`

```{r dt-annual}
annual |>
  mutate(
    registered_vehicles    = comma(registered_vehicles),
    total_accidents        = comma(total_accidents),
    fatal_injury_accidents = comma(fatal_injury_accidents),
    deaths_total           = comma(deaths_total),
    injured                = comma(injured),
    fatal_injury_share     = percent(fatal_injury_share, 0.1),
    deaths_per_accident    = round(deaths_per_accident, 4)
  ) |>
  datatable(
    rownames = FALSE,
    options  = list(pageLength = 10, dom = "t"),
    caption  = "Annual headline series, 2015–2024"
  )
```

### Interactive preview — `vehicles` (2024)

```{r dt-vehicles}
vehicles |>
  select(vehicle_type, vehicles_involved, vehicles_pct,
         driver_deaths_total, deaths_pct, injured_pct,
         deaths_per_1000_involved) |>
  arrange(desc(vehicles_involved)) |>
  mutate(deaths_per_1000_involved = round(deaths_per_1000_involved, 1)) |>
  datatable(
    rownames = FALSE,
    filter   = "top",
    options  = list(pageLength = 10, dom = "ftp"),
    caption  = "Vehicle-type involvement and driver outcomes, 2024"
  )
```

---

## Data Quality Notes

- **Coverage difference.** The annual TÜİK series covers *all* road accidents in
  Türkiye. Monthly/weekday/daylight/fault tables come from the KGM-responsibility
  network only (~20% of national fatal/injury accidents but ~53% of fatalities).
- **30-day fatality definition.** A person is counted as a road death if they die
  within 30 days of the accident from related injuries.
- **Missing 2024 vehicle-km.** Annual vkm for 2024 had not been published at
  collection time, so `rates$deaths_per_billion_vkm` and
  `rates$injured_per_billion_vkm` are `NA` for 2024.
