---
title: "Data"
subtitle: "Sources, preprocessing pipeline, and the unified dataset"
---

```{r setup, include=FALSE}
library(dplyr)
library(knitr)
library(scales)
library(DT)
load("data/housing_market_tr.RData")
```

## Data Sources

We integrate four official Turkish data sources into a unified analytical dataset:

| # | Source | Series | Coverage |
|---|---|---|---|
| 1 | **TCMB** — Konut Fiyat Endeksi | Nominal & real KFE, NUTS-1 regional KFE | Monthly, 2010–2024 |
| 2 | **TCMB EVDS** — Consumer Prices | CPI index for deflation | Monthly, 2010–2024 |
| 3 | **TCMB EVDS** — Exchange Rates | USD/TRY monthly average | Monthly, 2010–2024 |
| 4 | **TÜİK** — İnşaat ve Konut İstatistikleri | Housing sales counts, building permits | Annual, 2010–2024 |

**Direct links:**

- TCMB KFE: <https://www.tcmb.gov.tr/wps/wcm/connect/tr/tcmb+tr/main+menu/istatistikler/reel+sektor+istatistikleri/konut+fiyat+endeksi/>
- TCMB EVDS (CPI, Kur): <https://evds2.tcmb.gov.tr>
- TÜİK Konut İstatistikleri: <https://data.tuik.gov.tr/Kategori/GetKategori?p=insaat-ve-konut-116>

::: {.callout-note}
**Why this dataset?** Turkey's housing market has undergone one of the world's most
extreme price cycles in 2020–2024, driven by a unique combination of currency collapse,
triple-digit inflation, post-earthquake displacement, and supply-side contraction.
This makes it an exceptionally rich case study for understanding how macroeconomic
shocks transmit into asset markets — directly relevant to Turkey and actionable for
policy, investment, and academic audiences.
:::

---

## Preprocessing Pipeline

Raw TCMB and TÜİK data require aggregation and cleaning before analysis.
The full pipeline is in `R/01_preprocess.R`. Key steps:

**Step 1 — Manual extraction.**  
Annual and quarterly series were extracted from TCMB publications and TÜİK statistical
tables into `data/raw_*.csv` files, preserving original source values.

**Step 2 — Tidying and harmonisation.**  
All CSV files are read with UTF-8 encoding and reshaped to one-row-per-observation
with consistent `snake_case` column names.

**Step 3 — Derived variables.**

| Variable | Object | Formula |
|---|---|---|
| `kfe_nominal_yoy` | `kfe_annual` | (kfe_nominal / lag(kfe_nominal) − 1) × 100 |
| `kfe_real_yoy` | `kfe_annual` | (kfe_real / lag(kfe_real) − 1) × 100 |
| `cpi_yoy` | `kfe_annual` | (cpi_index / lag(cpi_index) − 1) × 100 |
| `cumulative_2020_2024_pct` | `regional_kfe` | Compounded 5-year nominal change |
| `real_premium` | `exchange_housing` | kfe_yoy_pct − cpi_yoy_pct |
| `second_hand_share` | `sales_permits` | sales_second_hand / sales_total × 100 |
| `mortgage_share` | `sales_permits` | sales_mortgage / sales_total × 100 |

```{r preprocess-show, eval=FALSE}
# R/01_preprocess.R  —  run once before rendering
library(readr); library(dplyr)

read_raw <- function(f) read_csv(file.path("data", f),
                                  show_col_types = FALSE,
                                  locale = locale(encoding = "UTF-8"))

kfe_annual <- read_raw("raw_kfe_annual.csv") |>
  mutate(
    kfe_nominal_yoy = (kfe_nominal / lag(kfe_nominal) - 1) * 100,
    kfe_real_yoy    = (kfe_real    / lag(kfe_real)    - 1) * 100,
    cpi_yoy         = (cpi_index   / lag(cpi_index)   - 1) * 100,
    kfe_cumulative  = kfe_nominal / 100
  )

regional_kfe <- read_raw("raw_regional_kfe.csv") |>
  mutate(
    avg_annual_pct = (kfe_2020_pct + kfe_2021_pct + kfe_2022_pct +
                      kfe_2023_pct + kfe_2024_pct) / 5,
    cumulative_2020_2024_pct =
      ((1 + kfe_2020_pct/100) * (1 + kfe_2021_pct/100) *
       (1 + kfe_2022_pct/100) * (1 + kfe_2023_pct/100) *
       (1 + kfe_2024_pct/100) - 1) * 100
  )

affordability    <- read_raw("raw_affordability.csv")
exchange_housing <- read_raw("raw_exchange_housing.csv") |>
  mutate(period = paste0(year, " ", quarter),
         real_premium = kfe_yoy_pct - cpi_yoy_pct)
sales_permits    <- read_raw("raw_sales_permits.csv") |>
  mutate(second_hand_share = sales_second_hand / sales_total * 100,
         mortgage_share    = sales_mortgage    / sales_total * 100)

save(kfe_annual, regional_kfe, affordability,
     exchange_housing, sales_permits,
     file = "data/housing_market_tr.RData")
```

---

## The Final Dataset

**[⬇ Download housing_market_tr.RData](data/housing_market_tr.RData)**

```r
load("data/housing_market_tr.RData")  # 5 data frame yüklenir
```

### Contents

```{r contents}
tibble::tibble(
  Object = c("kfe_annual","regional_kfe","affordability",
             "exchange_housing","sales_permits"),
  Description = c(
    "Annual KFE (nominal & real), CPI, USD/TRY, sales, permits — 2010–2024",
    "NUTS-1 regional annual KFE change % and median unit prices — 2020–2024",
    "Housing affordability: m² price, min. wage, months-to-buy — 2010–2024",
    "Quarterly USD/TRY vs KFE for correlation analysis — 2018–2024",
    "Annual housing sales by type and building permits — 2010–2024"
  ),
  Rows = c(nrow(kfe_annual), nrow(regional_kfe), nrow(affordability),
           nrow(exchange_housing), nrow(sales_permits))
) |> kable()
```

### Preview — `kfe_annual`

```{r dt-kfe}
kfe_annual |>
  select(year, kfe_nominal, kfe_real, cpi_index, usdtry_avg,
         kfe_nominal_yoy, kfe_real_yoy) |>
  mutate(across(c(kfe_nominal, kfe_real, cpi_index), ~round(., 1)),
         across(c(kfe_nominal_yoy, kfe_real_yoy), ~round(., 1))) |>
  datatable(rownames = FALSE,
            options  = list(pageLength = 15, dom = "t"),
            caption  = "Annual KFE series, 2010–2024")
```

### Preview — `regional_kfe`

```{r dt-regional}
regional_kfe |>
  select(region_code, region_name, kfe_2022_pct, kfe_2023_pct,
         kfe_2024_pct, avg_annual_pct, cumulative_2020_2024_pct,
         median_price_2024_try_m2) |>
  mutate(across(c(avg_annual_pct, cumulative_2020_2024_pct), ~round(., 1)),
         median_price_2024_try_m2 = comma(median_price_2024_try_m2)) |>
  datatable(rownames = FALSE, filter = "top",
            options  = list(pageLength = 12, dom = "ftp"),
            caption  = "Regional KFE — annual change % and 2024 median unit price")
```

---

## Data Quality Notes

- **KFE methodology revision (July 2024).** TCMB revised the KFE calculation from a
  quarterly to a monthly hedonic regression model, rebasing to 2023=100. The annual
  series here has been converted to a consistent 2010=100 base for comparability.

- **Regional data.** NUTS-1 regional breakdowns are published as annual % change
  figures. Median unit prices are TCMB-reported quarterly figures averaged to annual.

- **Sales data.** First-hand / second-hand split is available from 2015 onwards.
  Foreign buyer sales figures from TÜİK annual tables.

- **CPI deflation.** Real KFE is computed using TURKSTAT's official CPI index
  (2003=100 base, consistent series).
