---
title: "Road Traffic Accidents in Türkiye"
subtitle: "2015–2024 · EMU430 Data Analytics Term Project"
author:
  - name: "Arda Tüylüoğlu"
    url: https://github.com/KULLANICI_ADI_ARDA
  - name: "İlhan Emre Can"
    url: https://github.com/KULLANICI_ADI_ILHAN
  - name: "Kayra Yaltırık"
    url: https://github.com/KULLANICI_ADI_KAYRA
date: today
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(scales)
library(knitr)
library(tidyr)
library(DT)

load("data/traffic_accidents_tr.RData")

# Renk paleti — list olarak tanımla, pal$accent şeklinde eriş
pal <- list(
  primary   = "#1a1814",
  secondary = "#2c3e50",
  accent    = "#c0392b",
  good      = "#27ae60",
  warn      = "#e67e22",
  muted     = "#95a5a6"
)

# ggplot teması
theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title       = element_text(face = "bold", size = 13, hjust = 0),
      plot.subtitle    = element_text(colour = "#6b6560", size = 10.5, hjust = 0),
      plot.caption     = element_text(colour = "#95a5a6", size = 8, hjust = 0),
      plot.background  = element_rect(fill = "#f8f5f0", colour = NA),
      panel.background = element_rect(fill = "#f8f5f0", colour = NA),
      panel.grid.major = element_line(colour = "#e0dbd3"),
      panel.grid.minor = element_blank(),
      legend.position  = "top",
      legend.text      = element_text(size = 9),
      axis.text        = element_text(colour = "#6b6560", size = 9),
      axis.title       = element_text(colour = "#1a1814", size = 10)
    )
)
```

## Project Overview

Road traffic safety is one of Türkiye's most persistent public challenges. Between
2015 and 2024 the registered vehicle fleet grew by **57%** — yet fatal-or-injurious
accidents rose **46%**, reaching **266,854** in 2024. This project analyses a decade
of official accident statistics from **TÜİK**, **KGM**, and **EGM** to answer four
core questions:

1. How have fatality and injury rates evolved relative to fleet growth?
2. Where is the safety burden concentrated — seasonally, by vehicle type, by geography?
3. Which driver behaviours and crash types are most lethal?
4. How does Türkiye compare with European peers?

Explore the full analysis on the [Analysis](analysis.qmd) page and dataset details
on the [Data](data.qmd) page.

---

## 2024 Snapshot

```{r topline-table}
y    <- annual |> filter(year == 2024)
prev <- annual |> filter(year == 2023)

tibble::tibble(
  Indicator = c(
    "Registered vehicles",
    "Total accidents",
    "Fatal / injury accidents",
    "Deaths (within 30 days)",
    "Injured"
  ),
  `2024` = c(
    comma(y$registered_vehicles),
    comma(y$total_accidents),
    comma(y$fatal_injury_accidents),
    comma(y$deaths_total),
    comma(y$injured)
  ),
  `Change vs 2023` = c(
    percent((y$registered_vehicles    - prev$registered_vehicles)    / prev$registered_vehicles,    0.1),
    percent((y$total_accidents         - prev$total_accidents)         / prev$total_accidents,         0.1),
    percent((y$fatal_injury_accidents  - prev$fatal_injury_accidents)  / prev$fatal_injury_accidents,  0.1),
    percent((y$deaths_total            - prev$deaths_total)            / prev$deaths_total,            0.1),
    percent((y$injured                 - prev$injured)                 / prev$injured,                 0.1)
  )
) |>
  kable(align = "lrr")
```

---

## Key Takeaways

::: {.callout-note icon="false"}
### Summary of Findings

- **Volume up, per-vehicle risk down.** Total accidents grew 46% from 2015 to 2024,
  yet deaths per 100,000 registered vehicles fell from **37.7 to 20.3** — a ~46%
  reduction in vehicle-normalised fatality risk.

- **2020 was a pandemic-era anomaly.** Both accident volume and deaths fell sharply
  (deaths: 4,866 — decade low), then overshot pre-pandemic levels by 2023.

- **Driver fault dominates.** Driver errors account for ~88–90% of attributable
  causes every year. The single most common error in 2024 is *failing to adapt
  speed to road, weather, and traffic conditions* (40.8% of driver faults).

- **Motorcycles carry outsized risk.** They make up 31.5% of vehicles in
  fatal/injury accidents yet account for **40.0%** of driver deaths and **54.1%**
  of driver injuries.

- **Türkiye lags Europe.** With ~74 deaths per million population in 2024,
  Türkiye sits well above the EU-27 average of ~47.
:::
