---
title: "Turkish Housing Market"
subtitle: "Nominal Growth, Real Losses, and Affordability 2010-2024"
author:
  - name: "Arda Tuyluoglu"
    url: https://emu-hacettepe-analytics.github.io/emu430-spring2026-ardatuyluoglulab/
  - name: "Ilhan Emre Can"
    url: https://emu-hacettepe-analytics.github.io/emu430-spring2026-ilhanemrecan/
  - name: "Kayra Yaltirik"
    url: https://emu-hacettepe-analytics.github.io/emu430-spring2026-kayrayltrk/
date: today
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(scales)
library(knitr)
library(tidyr)
library(DT)

load("data/housing_market_tr.RData")

pal <- list(
  primary   = "#1a1814",
  secondary = "#2c3e50",
  accent    = "#c0392b",
  good      = "#27ae60",
  warn      = "#e67e22",
  muted     = "#95a5a6"
)

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title       = element_text(face = "bold", size = 13, hjust = 0),
      plot.subtitle    = element_text(colour = "#6b6560", size = 10.5, hjust = 0),
      plot.caption     = element_text(colour = "#95a5a6", size = 8, hjust = 0),
      plot.background  = element_rect(fill = "#f8f5f0", colour = NA),
      panel.background = element_rect(fill = "#f8f5f0", colour = NA),
      panel.grid.major = element_line(colour = "#e0dbd3"),
      panel.grid.minor = element_blank(),
      legend.position  = "top",
      legend.text      = element_text(size = 9),
      axis.text        = element_text(colour = "#6b6560", size = 9),
      axis.title       = element_text(colour = "#1a1814", size = 10)
    )
)
```

## Project Overview

Turkey has experienced one of the world's most dramatic housing price surges in recent
years, driven by high inflation, sharp currency depreciation, post-earthquake demand
shocks, and structural supply constraints. Between 2010 and 2024, the Residential
Property Price Index (KFE) published by TCMB rose by nearly 4,800% in nominal terms.
After adjusting for inflation, the real gain is far more modest and has recently
turned negative.

This project investigates the Turkish housing market through four questions:

1. How have housing prices evolved nominally vs. in real (inflation-adjusted) terms
   since 2010?
2. Which regions have experienced the most extreme price growth?
3. How has housing affordability changed, measured by the m2 price-to-minimum wage
   ratio?
4. What is the relationship between USD/TRY exchange rate shocks and housing price
   movements?

Explore the full analysis on the [Analysis](analysis.qmd) page and dataset details
on the [Data](data.qmd) page.

---

## 2024 Snapshot

```{r topline}
latest  <- annual_macro |> filter(year == 2024)
prev    <- annual_macro |> filter(year == 2023)
aff_now <- affordability |> filter(year == 2024)

tibble::tibble(
  Indicator = c(
    "Nominal KFE (2010 = 100)",
    "Real KFE (2010 = 100)",
    "Nominal KFE YoY change",
    "Real KFE YoY change",
    "Average m2 price (TRY)",
    "Average m2 price (USD)",
    "Months of min. wage for 100 m2"
  ),
  Value = c(
    comma(round(latest$kfe_nominal_2010base, 1)),
    comma(round(latest$kfe_real_2010base, 1)),
    paste0(round(latest$kfe_nominal_yoy, 1), "%"),
    paste0(round(latest$kfe_real_yoy, 1), "%"),
    paste0("TRY ", comma(aff_now$avg_m2_price_try)),
    paste0("USD ", comma(aff_now$avg_m2_price_usd)),
    round(aff_now$months_100m2, 0)
  )
) |>
  kable(align = "lr")
```

---

## Key Takeaways

::: {.callout-note icon="false"}
### Summary of Findings

- **Nominal surge, real stagnation.** KFE rose ~4,800% in nominal terms from 2010 to
  2024. After adjusting for CPI, the real index peaked in 2022 at 153.7 then fell back
  to 118.4 by 2024. Buyers largely paid an inflation premium, not a true wealth gain.

- **Affordability crisis.** In 2010, a 100 m2 flat cost ~110 monthly minimum wages.
  By 2022 this rose to ~276 months -- a 2.5x worsening driven by prices outpacing wages.

- **Exchange rate pass-through.** The 2018 currency crisis and the 2021-22 depreciation
  both preceded the sharpest nominal KFE increases. Quarterly FX and KFE changes are
  strongly correlated.

- **Istanbul and Aegean lead, Southeast lags.** Five-year cumulative nominal KFE gain
  (2020-2024) ranges from ~2,100% in the Aegean to ~1,400% in Southeast Anatolia.

- **Supply is in structural decline.** Building permits fell from 1.22 million units
  in 2015 to 549,000 in 2024 -- a 55% drop that signals undersupply ahead.
:::
