---
title: "Turkish Housing Market"
subtitle: "Nominal Growth, Real Losses, and Affordability 2010-2024"
author:
  - name: "Arda Tuyluoglu"
    url: https://github.com/KULLANICI_ADI_ARDA
  - name: "Ilhan Emre Can"
    url: https://github.com/KULLANICI_ADI_ILHAN
  - name: "Kayra Yaltirik"
    url: https://github.com/KULLANICI_ADI_KAYRA
date: today
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(scales)
library(knitr)
library(tidyr)
library(DT)

load("data/housing_market_tr.RData")

pal <- list(
  primary   = "#1a1814",
  secondary = "#2c3e50",
  accent    = "#c0392b",
  good      = "#27ae60",
  warn      = "#e67e22",
  muted     = "#95a5a6"
)

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title       = element_text(face = "bold", size = 13, hjust = 0),
      plot.subtitle    = element_text(colour = "#6b6560", size = 10.5, hjust = 0),
      plot.caption     = element_text(colour = "#95a5a6", size = 8, hjust = 0),
      plot.background  = element_rect(fill = "#f8f5f0", colour = NA),
      panel.background = element_rect(fill = "#f8f5f0", colour = NA),
      panel.grid.major = element_line(colour = "#e0dbd3"),
      panel.grid.minor = element_blank(),
      legend.position  = "top",
      legend.text      = element_text(size = 9),
      axis.text        = element_text(colour = "#6b6560", size = 9),
      axis.title       = element_text(colour = "#1a1814", size = 10)
    )
)
```

## Project Overview

Turkey has experienced one of the world's most dramatic housing price surges
in recent years, driven by high inflation, sharp currency depreciation,
post-earthquake demand shocks, and structural supply constraints.
Between 2010 and 2024, the Residential Property Price Index (KFE) published
by TCMB rose by nearly 4,800% in nominal terms. After adjusting for
inflation, the real gain is far more modest and has recently turned negative.

This project investigates the Turkish housing market through four questions:

1. How have housing prices evolved nominally vs. in real (inflation-adjusted)
   terms since 2010?
2. Which regions have experienced the most extreme price growth?
3. How has housing affordability changed, measured by the m2 price-to-minimum
   wage ratio?
4. What is the relationship between USD/TRY exchange rate shocks and housing
   price movements?

Explore the full analysis on the [Analysis](analysis.qmd) page and dataset
details on the [Data](data.qmd) page.

---

## 2024 Snapshot

```{r topline}
latest  <- kfe_annual |> filter(year == 2024)
prev    <- kfe_annual |> filter(year == 2023)
aff_now <- affordability |> filter(year == 2024)

tibble::tibble(
  Indicator = c(
    "Nominal KFE (2010 = 100)",
    "Real KFE (2010 = 100)",
    "Nominal KFE YoY change",
    "Real KFE YoY change",
    "Average m2 price (TRY)",
    "Average m2 price (USD)",
    "Months of min. wage for 100 m2"
  ),
  Value = c(
    comma(round(latest$kfe_nominal, 1)),
    comma(round(latest$kfe_real, 1)),
    paste0(round(latest$kfe_nominal_yoy, 1), "%"),
    paste0(round(latest$kfe_real_yoy, 1), "%"),
    paste0("TRY ", comma(aff_now$avg_m2_price_try)),
    paste0("USD ", comma(aff_now$avg_m2_price_usd)),
    round(aff_now$affordability_months, 0)
  )
) |>
  kable(align = "lr")
```

---

## Key Takeaways

::: {.callout-note icon="false"}
### Summary of Findings

- **Nominal surge, real stagnation.** KFE rose ~4,800% in nominal terms from
  2010 to 2024. After adjusting for CPI, the real index peaked in 2022 at
  153.7 then fell back to 118.4 by 2024. Buyers largely paid an inflation
  premium, not a true wealth gain.

- **Affordability crisis.** In 2010, a 100 m2 flat cost ~103 monthly minimum
  wages. By 2023 this rose to ~330 months -- a 3x worsening in 13 years.

- **Exchange rate pass-through.** The 2018 currency crisis (+60% USD/TRY) and
  the 2021-22 depreciation (+87%) both preceded the sharpest nominal KFE
  increases. Correlation between quarterly FX and KFE changes exceeds 0.70.

- **Istanbul and Aegean lead, Southeast lags.** Five-year cumulative nominal
  KFE gain (2020-2024) ranges from ~2,070% in the Aegean to ~1,420% in
  Southeast Anatolia.

- **Supply is in structural decline.** Building permits fell from 1.22 million
  units in 2015 to 549,000 in 2024 -- a 55% drop that signals undersupply
  through at least 2026-2027.
:::
