---
title: "Turkish Housing Market"
subtitle: "Nominal Growth, Real Losses, and Affordability — 2010–2024"
author:
  - name: "Arda Tüylüoğlu"
    url: https://github.com/KULLANICI_ADI_ARDA
  - name: "İlhan Emre Can"
    url: https://github.com/KULLANICI_ADI_ILHAN
  - name: "Kayra Yaltırık"
    url: https://github.com/KULLANICI_ADI_KAYRA
date: today
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(scales)
library(knitr)
library(tidyr)
library(DT)

load("data/housing_market_tr.RData")

pal <- list(
  primary   = "#1a1814",
  secondary = "#2c3e50",
  accent    = "#c0392b",
  good      = "#27ae60",
  warn      = "#e67e22",
  muted     = "#95a5a6"
)

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title       = element_text(face = "bold", size = 13, hjust = 0),
      plot.subtitle    = element_text(colour = "#6b6560", size = 10.5, hjust = 0),
      plot.caption     = element_text(colour = "#95a5a6", size = 8, hjust = 0),
      plot.background  = element_rect(fill = "#f8f5f0", colour = NA),
      panel.background = element_rect(fill = "#f8f5f0", colour = NA),
      panel.grid.major = element_line(colour = "#e0dbd3"),
      panel.grid.minor = element_blank(),
      legend.position  = "top",
      legend.text      = element_text(size = 9),
      axis.text        = element_text(colour = "#6b6560", size = 9),
      axis.title       = element_text(colour = "#1a1814", size = 10)
    )
)
```

## Project Overview

Turkey has experienced one of the world's most dramatic housing price surges in recent
years — driven by high inflation, sharp currency depreciation, post-earthquake demand
shocks, and structural supply constraints. Between 2010 and 2024, the **Residential
Property Price Index (KFE)** published by TCMB rose by nearly **4,800%** in nominal
terms. Yet once adjusted for inflation, the real gain is far more modest — and in
recent years has turned negative.

This project investigates the Turkish housing market through four lenses:

1. How have housing prices evolved **nominally vs. in real (inflation-adjusted) terms** since 2010?
2. Which **regions** have experienced the most extreme price growth, and what patterns emerge geographically?
3. How has **housing affordability** changed — measured by the m² price-to-minimum wage ratio?
4. What is the relationship between **USD/TRY exchange rate shocks** and housing price movements?

Explore the full analysis on the [Analysis](analysis.qmd) page and dataset details
on the [Data](data.qmd) page.

---

## 2024 Snapshot

```{r topline}
latest  <- kfe_annual |> filter(year == 2024)
prev    <- kfe_annual |> filter(year == 2023)
base    <- kfe_annual |> filter(year == 2010)
aff_now <- affordability |> filter(year == 2024)
aff_10  <- affordability |> filter(year == 2010)

tibble::tibble(
  Indicator = c(
    "Nominal KFE (2010 = 100)",
    "Real KFE (2010 = 100)",
    "Nominal KFE — YoY change",
    "Real KFE — YoY change",
    "Average m² price (TRY)",
    "Average m² price (USD)",
    "Months of min. wage for 100 m²"
  ),
  Value = c(
    comma(round(latest$kfe_nominal, 1)),
    comma(round(latest$kfe_real, 1)),
    paste0(round(latest$kfe_nominal_yoy, 1), "%"),
    paste0(round(latest$kfe_real_yoy, 1), "%"),
    paste0("₺", comma(aff_now$avg_m2_price_try)),
    paste0("$", comma(aff_now$avg_m2_price_usd)),
    round(aff_now$affordability_months, 0)
  )
) |>
  kable(align = "lr")
```

---

## Key Takeaways

::: {.callout-note icon="false"}
### Summary of Findings

- **Nominal surge, real stagnation.** The KFE rose ~4,800% in nominal terms from 2010
  to 2024. After adjusting for CPI, the real index peaked in **2022 at 153.7** then
  fell back — in 2024 real prices were below the 2021 peak, meaning buyers largely paid
  an inflation premium, not a true wealth gain.

- **The affordability crisis is structural.** In 2010, a 100 m² flat cost ~103 monthly
  minimum wages. By 2023 this had risen to **~330 months** — a 3× worsening. USD-priced
  affordability also deteriorated sharply as the lira depreciated.

- **Exchange rate shocks transmit rapidly.** The 2018 currency crisis (+60% USD/TRY in
  one year) and the 2021–22 depreciation episode (+87%) both preceded or coincided with
  the sharpest nominal KFE increases — suggesting a strong pass-through effect.

- **Istanbul and Aegean lead, Southeast lags.** The five-year (2020–2024) cumulative
  KFE increase ranged from ~21× in the Aegean to ~14× in Southeast Anatolia — a 1.5×
  geographic spread in nominal terms.

- **Supply is contracting.** Building permits peaked at 1.2 million units in 2015 and
  fell to ~549,000 in 2024. This structural supply decline, combined with post-earthquake
  displacement demand, is reinforcing price pressure.
:::
